---
title: "Getting started with edaphos"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Getting started with edaphos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4,
  out.width  = "90%"
)
```

```{r setup}
library(edaphos)
set.seed(1L)
```

## What this vignette is for

edaphos is organised around **ten research pillars**.  This vignette
is the 200-line orientation tour: one minimal, self-contained
example per pillar and per cross-pillar bridge, all running on
synthetic Cerrado-like data that ships with the package.

Every function called here runs offline in seconds.  For the full
treatment of each pillar -- real WoSIS datasets, MCMC diagnostics,
head-to-head benchmarks -- see the per-pillar vignettes in the
sidebar.

## A shared toy dataset

A 60-profile synthetic data frame with pedological covariates and
topsoil soil-organic-carbon (SOC) as the regression target.  Every
pillar below consumes this same object.

```{r toy-data}
n <- 60L
dat <- data.frame(
  lon   = runif(n, -50, -48),
  lat   = runif(n, -16, -14),
  map   = runif(n, 900, 1500),   # mean annual precipitation
  mat   = runif(n, 21, 27),      # mean annual temperature
  clay  = runif(n, 10, 50),
  sand  = runif(n, 20, 70),
  trees = runif(n, 0, 50)
)
dat$soc <- with(dat,
  8 + 0.02 * map + 0.10 * clay - 0.15 * trees + rnorm(n, 0, 2))
head(dat)
```

Train / test split we will reuse throughout:

```{r split}
tr_ix <- sample.int(n, 45L)
tr    <- dat[tr_ix,  ]
te    <- dat[-tr_ix, ]
cov_cols <- c("map", "mat", "clay", "sand", "trees")
```

## Pilar 1 -- Causal AI with backdoor adjustment

A pedogenetic DAG gives the minimally sufficient set of adjustment
variables; `benchmark_fit_p1_causal()` then runs a bootstrap OLS
restricted to that set and produces a predictive posterior.

```{r p1}
post_p1 <- benchmark_fit_p1_causal(tr, te, cov_cols,
                                       dag = NULL, n_boot = 100L)
uncertainty_calibrate(post_p1, truth = te$soc)$crps
```

## Pilar 2 -- Physics-Informed profile ODE

The pedogenetic ODE `dy/dz = -lambda0 exp(-mu z) (y - y_inf)` is
fit to a single pedon's SOC profile with `piml_profile_fit()`.

```{r p2}
depths <- c(5, 15, 30, 60, 100)
values <- c(25, 18, 12, 8, 6.5)
ode_fit <- piml_profile_fit(depths, values)
ode_fit$params
```

## Pilar 3 -- Temporal 4D cube (ConvLSTM)

```{r p3, eval = FALSE}
# Skipped in the getting-started tour -- ConvLSTM training needs
# a temporal raster stack, covered in `pilar3-4d-soc.Rmd`.
```

## Pilar 4 -- Foundation models (SimCLR / MoCo)

```{r p4, eval = FALSE}
# Pre-trained MoCo weights live on Zenodo and are pulled by
# `foundation_weights_load()`; see `pilar4-simclr-embeddings.Rmd`.
```

## Pilar 5 -- Active Learning

A hybrid uncertainty + diversity acquisition function selects the
next 3 profiles to sample from a pool of candidates.

```{r p5}
al_mod <- al_fit(labeled    = tr,
                   target     = "soc",
                   covariates = cov_cols,
                   coords     = c("lon", "lat"),
                   num.trees  = 200L)
pool <- dat[sample.int(n, 20L), c(cov_cols, "lon", "lat")]
picks <- al_query(al_mod, pool, n = 3L, strategy = "hybrid")
picks
```

## Pilar 6 -- Quantum KRR

A 6-qubit ZZFeatureMap kernel ridge regression trained on the PCA
of the covariates, with a bootstrap-ensemble posterior.

```{r p6}
post_p6 <- benchmark_fit_p6_quantum(tr, te, cov_cols,
                                         n_pcs = 4L, reps = 1L,
                                         n_boot = 5L, lambda = 0.5)
uncertainty_calibrate(post_p6, truth = te$soc)$crps
```

## Pilar 7 -- Bayesian Hierarchical Spatial

```{r p7}
fit7 <- bhs_fit(tr, soc ~ map + mat + clay + sand + trees,
                  coords = c("lon", "lat"),
                  nmcmc = 200L, burn = 100L,
                  phi_range = c(0.1, 5), verbose = FALSE)
pr7 <- predict(fit7, te, n_draws = 80L)
head(pr7[, c("mean", "sd")])
```

## Pilar 8 -- Neural operators

```{r p8}
depth_grid <- seq(5, 120, length.out = 8L)
targets   <- t(apply(tr[, cov_cols], 1L, function(r)
  10 + 5 * r[1] + 2 * runif(length(depth_grid))))
no_fit <- no_deeponet_fit(depth_grid, targets,
                              as.matrix(tr[, cov_cols]),
                              epochs = 50L, seed = 1L)
no_pred <- predict(no_fit, as.matrix(te[, cov_cols]))
dim(no_pred)
```

## Pilar 9 -- Diffusion models

A conditional DDPM trained on synthetic 6x6 patches; `dm_sample()`
draws 4 posterior maps.

```{r p9}
patches <- array(rnorm(16L * 6L * 6L), dim = c(16L, 6L, 6L))
dm <- dm_fit(patches, T = 8L, epochs = 10L,
               hidden = 8L, lr = 0.05, seed = 1L)
samps <- dm_sample(dm, n_samples = 4L, seed = 1L)
dim(samps)
```

## Pilar 10 -- Graph Attention Network

A k-NN co-location graph on (lon, lat) with covariates as node
features, fit via `gnn_fit()`.

```{r p10}
g <- gnn_build_graph(dat, k = 5L, feature_cols = cov_cols)
gat <- gnn_fit(g, targets = dat$soc,
                 hidden = 8L, n_heads = 2L, n_layers = 1L,
                 epochs = 30L, lr = 0.03, seed = 1L)
gat$history[c(1, 30)]
```

## Cross-pillar bridges (v3.0.0)

The **bridges** compose two pillars into a single API.  Here we
show a representative pair per cluster.

### P10 x P1 -- GAT embeddings as causal nuisance

```{r bridge-p10-p1, eval = FALSE}
# Requires `bnlearn`; run when that Suggests dependency is installed.
feat_df <- dat[, c(cov_cols, "soc")]
kg <- gnn_causal_discovery(gat, feat_df, method = "hc",
                              n_emb_cols = 3L, seed = 1L)
head(kg$edges_feature_only)
```

### P7 x P5 -- Thompson-sampling AL

Posterior-sampling active learning over the BHS fit.

```{r bridge-p7-p5}
pool_df <- dat[1:8, c(cov_cols, "lon", "lat")]
q <- al_query_bhs(fit7, pool_df, n_select = 3L, n_draws = 50L)
q
```

## Head-to-head: unified uncertainty scoring

All six methods producing `edaphos_posterior` objects can be
scored side-by-side through `uncertainty_calibrate()`.

```{r head-to-head}
crps <- c(
  P1_Causal   = uncertainty_calibrate(post_p1, te$soc)$crps,
  P6_Quantum  = uncertainty_calibrate(post_p6, te$soc)$crps
)
crps
```

For the full 1 095-profile WoSIS Cerrado benchmark comparing all
six "static point regression" methods, load
`inst/extdata/benchmark_wosis_6pilar.rds` and inspect
`cv_aggregate` (v3.1.0 vignette: `case-cerrado-end-to-end.Rmd`).

## Next steps

* **Per-pillar deep dives** -- 17 vignettes, one per pillar + key
  case studies, cover the full feature surface.
* **`?edaphos`** -- package-level help page with the complete
  function index.
* **Cross-pillar bridges** -- see `causal-discovery-trio.Rmd` for
  how P1, P5 and P7 combine to run a DAG-driven AL loop.

## Citation

Please cite edaphos as:

> Rodrigues, H. (2026).  *edaphos: Disruptive algorithms for digital
> soil mapping.*  R package version `r as.character(utils::packageVersion("edaphos"))`.
> <https://github.com/HugoMachadoRodrigues/edaphos>
