---
title: "Pilar 6 — Quantum Kernel Methods for High-Dimensional Soil Covariates"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 6 — Quantum Kernel Methods for High-Dimensional Soil Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 5
)
```

## Abstract

High-dimensional covariate stacks — a natural product of modern Digital
Soil Mapping pipelines [@McBratney2003; @Wadoux2020] — eventually hit
the classical curse of dimensionality: the volume of the covariate
hypercube grows exponentially with the number of features, while the
number of labelled soil samples remains stubbornly linear in budget
[@Brus2019]. The **Pillar 6** of `edaphos` addresses this gap by
reformulating the kernel of the regression problem as the *overlap of
parametrised quantum states* [@Havlicek2019; @Schuld2019qml]. The
resulting **quantum kernel** lives in a complex Hilbert space whose
dimension grows as $2^n$ in the number of covariates $n$, yet remains
tractable to simulate in pure R for $n \le 8$ and is the subject of an
active research program on real NISQ hardware
[@Preskill2018nisq; @Biamonte2017qml].

The implementation in this release is a **self-contained, dependency-
free pure-R state-vector simulator** of the ZZFeatureMap circuit of
[@Havlicek2019], a quantum kernel Gram matrix evaluator, and a closed-
form Kernel Ridge Regression (KRR) wrapper. We demonstrate the full
pipeline on a binary NDVI-classification task drawn from the bundled
`br_cerrado` dataset, using the three covariates that drive NDVI in
the data-generating process.

## 1. Background: the curse of dimensionality as a Hilbert-space question

Classical DSM regressors (Random Forests, GAMs, Gaussian Processes)
live on the data itself: features enter through a (possibly non-linear)
kernel $k(\mathbf{x}_i, \mathbf{x}_j)$ on $\mathbb{R}^n$. The RBF
kernel $k_{\text{RBF}}(\mathbf{x}_i, \mathbf{x}_j) = \exp(-\gamma
\lVert\mathbf{x}_i - \mathbf{x}_j\rVert^2)$ is equivalent to an inner
product in an infinite-dimensional feature space, but this space is
*classically computable* and its expressivity is bounded by the
behaviour of $\gamma$.

A **quantum feature map** lifts each data point $\mathbf{x}$ to a
quantum state $\lvert\phi(\mathbf{x})\rangle$ in the Hilbert space of
$n$ qubits, $\mathcal{H} = \mathbb{C}^{2^n}$. The induced **quantum
kernel** is the squared overlap of those states:

$$
K(\mathbf{x}_i, \mathbf{x}_j) \;=\;
  \bigl\lvert\langle \phi(\mathbf{x}_j) \mid \phi(\mathbf{x}_i)\rangle\bigr\rvert^{2}.
$$

For a carefully designed feature map, computing $K$ on a classical
computer is believed to be **exponentially hard**
[@Havlicek2019, Supplementary Information] — the whole point of the
quantum hardware is to evaluate overlaps in that space directly. For
the small-$n$ problems that actually arise in DSM covariate stacks
(typical $n = 4\text{–}8$), however, classical simulation is cheap,
and the resulting kernel is a *new scientific object* to study even
without a quantum device.

## 2. The ZZFeatureMap encoding

The Havlicek et al. feature map constructs
$\lvert\phi(\mathbf{x})\rangle$ from a parameterised unitary
$U_\phi(\mathbf{x})$ applied to a uniform superposition:

$$
\lvert\phi(\mathbf{x})\rangle
   \;=\; \bigl(U_\phi(\mathbf{x})\,H^{\otimes n}\bigr)^{R}\,\lvert 0\rangle^{\otimes n},
$$

with $R \ge 1$ the number of encoding repetitions and

$$
U_\phi(\mathbf{x})
  \;=\; \exp\!\Bigl(i\sum_{S\subseteq[n]}\phi_S(\mathbf{x})\prod_{i\in S} Z_i\Bigr).
$$

In the second-order regime the non-trivial subsets are the singletons
$\{i\}$ and the pairs $\{i, j\}$, for which we take

$$
\phi_{\{i\}}(\mathbf{x}) \;=\; 2\,x_i,
\qquad
\phi_{\{i, j\}}(\mathbf{x}) \;=\; 2\,(\pi - x_i)(\pi - x_j).
$$

Pairwise terms are implemented via the standard
$\mathrm{CNOT}_{i\to j}\,R_z\bigl(\phi_{\{i,j\}}(\mathbf{x})\bigr)\,\mathrm{CNOT}_{i\to j}$
decomposition. This is the exact circuit that
[`.quantum_zz_feature_map()`][quantum_feature_map] implements in
`R/quantum_kernel.R`.

## 3. Implementation: a dependency-free R simulator

For up to 8 qubits the state vector has length $\le 256$, and every
gate can be applied by an $O(2^n)$ bit-level permutation of that
vector — no explicit $2^n \times 2^n$ Kronecker product is built. This
keeps `edaphos` free of any external quantum-simulator dependency
(no `qsimulatR`, no `reticulate` + PennyLane) for the core pathway,
while still reproducing the Havlicek *Nature* 2019 feature map
bit-for-bit.

```{r setup, message=FALSE}
library(edaphos)
set.seed(1)

# Sanity check: a 3-qubit state is normalised.
psi <- quantum_feature_map(c(pi / 4, pi / 3, pi / 2), reps = 2L)
length(psi)                 # 2^3 = 8 complex amplitudes
sum(Mod(psi)^2)             # <psi|psi> = 1
```

## 4. Quantum kernel on the bundled Cerrado dataset

We use the three `br_cerrado` covariates that **drive NDVI** in the
data-generating process — `slope`, `twi`, `map_mm` — and target the
binary median-split of NDVI itself. Three qubits give a Hilbert space
of dimension $2^3 = 8$. The features are rescaled into $[0, \pi]$ by
[`quantum_scale()`][quantum_scale], the de-facto range of validity for
the ZZFeatureMap.

```{r cerrado-data}
data(br_cerrado, package = "edaphos")

covs <- c("slope", "twi", "map_mm")    # three drivers of NDVI in the DGP
set.seed(1)
idx  <- sample(nrow(br_cerrado), 200L)
Xraw <- as.matrix(br_cerrado[idx, covs])
X    <- quantum_scale(Xraw)            # rescale columns -> [0, pi]

# Binary NDVI target: above / below the local median
y <- sign(br_cerrado$ndvi[idx] - stats::median(br_cerrado$ndvi[idx]))
y[y == 0] <- 1L
table(y)
```

```{r kernel-matrix}
K <- quantum_kernel(X, reps = 2L)
dim(K); round(K[1:5, 1:5], 3)
```

The Gram matrix is symmetric, with `K[i, i] = 1` by construction
(self-overlap of a normalised state) and all off-diagonal entries
in $[0, 1]$ — the numerical fingerprint of a valid quantum kernel.

```{r kernel-psd}
lambda_min <- min(eigen(K, symmetric = TRUE, only.values = TRUE)$values)
stopifnot(lambda_min > -1e-8)             # positive semi-definite
```

## 5. Quantum Kernel Ridge Regression on binary NDVI

With the Gram matrix in hand, Kernel Ridge Regression is a closed-form
problem:

$$
\boldsymbol{\alpha} \;=\; (K + \lambda I)^{-1}\,\mathbf{y}.
$$

[`quantum_krr_fit()`][quantum_krr_fit] computes this solve and
packages the training covariates, the dual coefficients
$\boldsymbol{\alpha}$ and the kernel configuration into a
`edaphos_quantum_krr` S3 object.

```{r qkrr-fit}
set.seed(1)
train <- sort(sample(length(y), 140L))
test  <- setdiff(seq_along(y), train)

fit <- quantum_krr_fit(
  X[train, ], y[train],
  reps   = 2L,
  lambda = 0.1
)
fit
```

```{r qkrr-predict}
# Numeric scores for soft evaluation
scores <- predict(fit, X[test, ], type = "numeric")
# Hard labels (pm 1) via sign(scores)
pred   <- predict(fit, X[test, ], type = "class")
table(predicted = pred, truth = y[test])
mean(pred == y[test])
```

This is not a benchmark — the point is the **scientific plumbing**:
every piece of the quantum-kernel pipeline is computable in a few
hundred lines of pure R, leaving the mathematical and physical
interpretation visible at every step.

## 6. Interpreting the kernel visually

```{r kernel-heatmap, fig.width=6, fig.height=6}
ord <- order(fit$fitted)
K_sorted <- fit$K_train[ord, ord]

image(seq_len(nrow(K_sorted)), seq_len(ncol(K_sorted)),
      K_sorted[, rev(seq_len(ncol(K_sorted)))],
      xlab = "train sample (ordered by predicted SOC score)",
      ylab = "train sample",
      main = "Quantum kernel — ZZFeatureMap on br_cerrado",
      col  = hcl.colors(64, "YlOrRd"),
      axes = FALSE)
```

Sorting the training samples by their predicted NDVI score reveals
diagonal blocks where the kernel recognises within-class similarity;
off-diagonal entries quantify across-class separation in the quantum
Hilbert space.

## 7. Variational Quantum Eigensolver for organo-mineral chemistry

Beyond the kernel methods of §1-6, Pillar 6 ships a second,
Qiskit-backed pipeline for **ground-state energy estimation** of
soil-relevant Hamiltonians via the Peruzzo et al. (2014)
**Variational Quantum Eigensolver** (VQE) [@Peruzzo2014vqe]. This is
the direct avenue to "simulation of materials" — predicting soil
fertility from the first-principles energy of clay-humus adsorption
or iron-oxide weathering complexes rather than from surface
correlations [@McClean2016vqe].

The R side exposes the entire Qiskit stack through four public
functions, all mediated by `reticulate`:

| Function | Purpose |
|---------|---------|
| `quantum_hamiltonian(pauli_terms)` | Wrap a named vector of Pauli coefficients into a `qiskit.quantum_info.SparsePauliOp`. |
| `quantum_hamiltonian_h2()` | Textbook 2-qubit H\(_2\) in the Bravyi-Kitaev-tapered STO-3G basis (the canonical VQE benchmark). |
| `quantum_hamiltonian_organo_mineral()` | 4-qubit toy Fe + ligand coupling (two metal-centre qubits + two ligand qubits, on-site / exchange / hopping terms). |
| `quantum_hamiltonian_ising_1d(n, J, h)` | Transverse-field Ising chain on `n` qubits. |
| `quantum_vqe_exact(ham)` | Exact ground-state via `NumPyMinimumEigensolver`. |
| `quantum_vqe_fit(ham, ansatz_reps, optimizer, max_iter, backend)` | Full VQE — EfficientSU2 ansatz + classical optimiser (COBYLA / SPSA / SLSQP / L-BFGS-B) + `StatevectorEstimator`; captures the energy trajectory via callback. |
| `quantum_ibmq_available()` / `quantum_ibmq_backends()` | Preflight + listing for real IBM-Q hardware via `qiskit-ibm-runtime` (optional). |

### 7.1 H\(_2\) as a reference

The H\(_2\) molecule at bond length 0.735 Å is the archetypal VQE
benchmark. Its Bravyi-Kitaev-tapered Hamiltonian has only five Pauli
terms on two qubits, and the ground-state energy is known to be
\(-1.857275\) Hartree.

```{r vqe-setup, eval=requireNamespace("reticulate", quietly = TRUE) && reticulate::py_module_available("qiskit")}
library(edaphos)

ham_h2 <- quantum_hamiltonian_h2()
ham_h2
```

```{r vqe-h2, eval=requireNamespace("reticulate", quietly = TRUE) && reticulate::py_module_available("qiskit")}
exact_h2 <- quantum_vqe_exact(ham_h2)
vqe_h2   <- quantum_vqe_fit(ham_h2, ansatz_reps = 2L,
                              optimizer = "COBYLA",
                              max_iter  = 200L,
                              seed = 1L)
vqe_h2
```

Typical COBYLA run: the VQE energy lands within \(10^{-5}\) Hartree
of the exact diagonalisation — the noise-free simulator reproduces
the published benchmark to machine precision.

### 7.2 Toy organo-mineral Hamiltonian

`quantum_hamiltonian_organo_mineral()` defines a deliberately small
4-qubit model that is nevertheless *structurally* similar to a
clay-humus adsorption complex:

\[
H = -\varepsilon_\mathrm{Fe}(Z_3 + Z_2)
    - \varepsilon_\mathrm{L}(Z_1 + Z_0)
    + J_\mathrm{Fe}\,Z_3 Z_2 + J_\mathrm{L}\,Z_1 Z_0
    + t\,(X_3 X_0 + X_2 X_1),
\]

with two metal-centre qubits (left pair) and two ligand qubits
(right pair). On-site energies, same-sector Z-Z exchange, and
cross-sector X-X hopping produce a non-trivial entangled ground state
that cannot be written as a product of the Fe and ligand wave-
functions — exactly the regime where a classical model-free regressor
struggles.

```{r vqe-organo, eval=requireNamespace("reticulate", quietly = TRUE) && reticulate::py_module_available("qiskit")}
ham_om <- quantum_hamiltonian_organo_mineral()
exact_om <- quantum_vqe_exact(ham_om)
vqe_om   <- quantum_vqe_fit(ham_om, ansatz_reps = 3L,
                              optimizer = "COBYLA",
                              max_iter  = 400L,
                              seed = 1L)
vqe_om
```

```{r vqe-curve, eval=requireNamespace("reticulate", quietly = TRUE) && reticulate::py_module_available("qiskit") && requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
df <- data.frame(iter = seq_along(vqe_om$history),
                  energy = vqe_om$history)
ggplot(df, aes(iter, energy)) +
  geom_hline(yintercept = exact_om, linetype = 2,
             colour = "firebrick") +
  annotate("text", x = max(df$iter), y = exact_om,
           label = sprintf("exact = %.4f", exact_om),
           hjust = 1, vjust = -0.5, colour = "firebrick") +
  geom_line(linewidth = 0.7, colour = "#7E1E9C") +
  labs(x = "COBYLA iteration",
       y = expression(paste("VQE energy estimate ", E[theta])),
       title = "Pillar 6 — VQE convergence on the organo-mineral Hamiltonian") +
  theme_minimal(base_size = 12)
```

### 7.3 Running on real IBM Quantum hardware (optional)

To run the same circuit on NISQ hardware [@Preskill2018nisq],

```r
install.packages("reticulate")
reticulate::py_install("qiskit-ibm-runtime", pip = TRUE)
Sys.setenv(IBMQ_TOKEN = "<your-ibm-quantum-api-token>")

quantum_ibmq_available()      # TRUE if everything is wired up
quantum_ibmq_backends()       # list the backends available to your account
```

The R-side API is unchanged; a future release will thread
`backend = "ibmq"` all the way through `quantum_vqe_fit()` by
swapping the local `StatevectorEstimator` for the
`qiskit_ibm_runtime.EstimatorV2` primitive.

## 8. Limits and roadmap

The pure-R simulator in §1-6 scales comfortably to **8 qubits**
(256-dim state vector; fractions of a second per kernel entry). The
Qiskit-backed VQE of §7 extends that to **~12 qubits** on a laptop
via the exact statevector simulator, and to **hundreds of qubits** on
real NISQ hardware through the `qiskit-ibm-runtime` bridge.

Three directions remain on the roadmap:

1. **Shot-based noisy simulation.** Wire the `StatevectorSampler` and
   Aer noise models through `quantum_vqe_fit(backend = "aer_sampler")`
   so users can stress-test the VQE against realistic readout and
   two-qubit gate error rates without leaving the laptop.
2. **Full IBM-Q dispatch.** Expose `backend = "ibmq"` end-to-end with
   automatic circuit transpilation and batched Runtime Primitives.
3. **Larger organo-mineral Hamiltonians.** Second-quantised fermionic
   Hamiltonians lifted through Jordan-Wigner / Bravyi-Kitaev, built
   on top of `qiskit-nature` active spaces — enough to approach
   realistic clay-humus coordination with useful accuracy.

## References
