---
title: "Pilar 6 — Quantum Kernel Methods for High-Dimensional Soil Covariates"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 6 — Quantum Kernel Methods for High-Dimensional Soil Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 5
)
```

## Abstract

High-dimensional covariate stacks — a natural product of modern Digital
Soil Mapping pipelines [@McBratney2003; @Wadoux2020] — eventually hit
the classical curse of dimensionality: the volume of the covariate
hypercube grows exponentially with the number of features, while the
number of labelled soil samples remains stubbornly linear in budget
[@Brus2019]. The **Pillar 6** of `edaphos` addresses this gap by
reformulating the kernel of the regression problem as the *overlap of
parametrised quantum states* [@Havlicek2019; @Schuld2019qml]. The
resulting **quantum kernel** lives in a complex Hilbert space whose
dimension grows as $2^n$ in the number of covariates $n$, yet remains
tractable to simulate in pure R for $n \le 8$ and is the subject of an
active research program on real NISQ hardware
[@Preskill2018nisq; @Biamonte2017qml].

The implementation in this release is a **self-contained, dependency-
free pure-R state-vector simulator** of the ZZFeatureMap circuit of
[@Havlicek2019], a quantum kernel Gram matrix evaluator, and a closed-
form Kernel Ridge Regression (KRR) wrapper. We demonstrate the full
pipeline on a binary NDVI-classification task drawn from the bundled
`br_cerrado` dataset, using the three covariates that drive NDVI in
the data-generating process.

## 1. Background: the curse of dimensionality as a Hilbert-space question

Classical DSM regressors (Random Forests, GAMs, Gaussian Processes)
live on the data itself: features enter through a (possibly non-linear)
kernel $k(\mathbf{x}_i, \mathbf{x}_j)$ on $\mathbb{R}^n$. The RBF
kernel $k_{\text{RBF}}(\mathbf{x}_i, \mathbf{x}_j) = \exp(-\gamma
\lVert\mathbf{x}_i - \mathbf{x}_j\rVert^2)$ is equivalent to an inner
product in an infinite-dimensional feature space, but this space is
*classically computable* and its expressivity is bounded by the
behaviour of $\gamma$.

A **quantum feature map** lifts each data point $\mathbf{x}$ to a
quantum state $\lvert\phi(\mathbf{x})\rangle$ in the Hilbert space of
$n$ qubits, $\mathcal{H} = \mathbb{C}^{2^n}$. The induced **quantum
kernel** is the squared overlap of those states:

$$
K(\mathbf{x}_i, \mathbf{x}_j) \;=\;
  \bigl\lvert\langle \phi(\mathbf{x}_j) \mid \phi(\mathbf{x}_i)\rangle\bigr\rvert^{2}.
$$

For a carefully designed feature map, computing $K$ on a classical
computer is believed to be **exponentially hard**
[@Havlicek2019, Supplementary Information] — the whole point of the
quantum hardware is to evaluate overlaps in that space directly. For
the small-$n$ problems that actually arise in DSM covariate stacks
(typical $n = 4\text{–}8$), however, classical simulation is cheap,
and the resulting kernel is a *new scientific object* to study even
without a quantum device.

## 2. The ZZFeatureMap encoding

The Havlicek et al. feature map constructs
$\lvert\phi(\mathbf{x})\rangle$ from a parameterised unitary
$U_\phi(\mathbf{x})$ applied to a uniform superposition:

$$
\lvert\phi(\mathbf{x})\rangle
   \;=\; \bigl(U_\phi(\mathbf{x})\,H^{\otimes n}\bigr)^{R}\,\lvert 0\rangle^{\otimes n},
$$

with $R \ge 1$ the number of encoding repetitions and

$$
U_\phi(\mathbf{x})
  \;=\; \exp\!\Bigl(i\sum_{S\subseteq[n]}\phi_S(\mathbf{x})\prod_{i\in S} Z_i\Bigr).
$$

In the second-order regime the non-trivial subsets are the singletons
$\{i\}$ and the pairs $\{i, j\}$, for which we take

$$
\phi_{\{i\}}(\mathbf{x}) \;=\; 2\,x_i,
\qquad
\phi_{\{i, j\}}(\mathbf{x}) \;=\; 2\,(\pi - x_i)(\pi - x_j).
$$

Pairwise terms are implemented via the standard
$\mathrm{CNOT}_{i\to j}\,R_z\bigl(\phi_{\{i,j\}}(\mathbf{x})\bigr)\,\mathrm{CNOT}_{i\to j}$
decomposition. This is the exact circuit that
[`.quantum_zz_feature_map()`][quantum_feature_map] implements in
`R/quantum_kernel.R`.

## 3. Implementation: a dependency-free R simulator

For up to 8 qubits the state vector has length $\le 256$, and every
gate can be applied by an $O(2^n)$ bit-level permutation of that
vector — no explicit $2^n \times 2^n$ Kronecker product is built. This
keeps `edaphos` free of any external quantum-simulator dependency
(no `qsimulatR`, no `reticulate` + PennyLane) for the core pathway,
while still reproducing the Havlicek *Nature* 2019 feature map
bit-for-bit.

```{r setup, message=FALSE}
library(edaphos)
set.seed(1)

# Sanity check: a 3-qubit state is normalised.
psi <- quantum_feature_map(c(pi / 4, pi / 3, pi / 2), reps = 2L)
length(psi)                 # 2^3 = 8 complex amplitudes
sum(Mod(psi)^2)             # <psi|psi> = 1
```

## 4. Quantum kernel on the bundled Cerrado dataset

We use the three `br_cerrado` covariates that **drive NDVI** in the
data-generating process — `slope`, `twi`, `map_mm` — and target the
binary median-split of NDVI itself. Three qubits give a Hilbert space
of dimension $2^3 = 8$. The features are rescaled into $[0, \pi]$ by
[`quantum_scale()`][quantum_scale], the de-facto range of validity for
the ZZFeatureMap.

```{r cerrado-data}
data(br_cerrado, package = "edaphos")

covs <- c("slope", "twi", "map_mm")    # three drivers of NDVI in the DGP
set.seed(1)
idx  <- sample(nrow(br_cerrado), 200L)
Xraw <- as.matrix(br_cerrado[idx, covs])
X    <- quantum_scale(Xraw)            # rescale columns -> [0, pi]

# Binary NDVI target: above / below the local median
y <- sign(br_cerrado$ndvi[idx] - stats::median(br_cerrado$ndvi[idx]))
y[y == 0] <- 1L
table(y)
```

```{r kernel-matrix}
K <- quantum_kernel(X, reps = 2L)
dim(K); round(K[1:5, 1:5], 3)
```

The Gram matrix is symmetric, with `K[i, i] = 1` by construction
(self-overlap of a normalised state) and all off-diagonal entries
in $[0, 1]$ — the numerical fingerprint of a valid quantum kernel.

```{r kernel-psd}
lambda_min <- min(eigen(K, symmetric = TRUE, only.values = TRUE)$values)
stopifnot(lambda_min > -1e-8)             # positive semi-definite
```

## 5. Quantum Kernel Ridge Regression on binary NDVI

With the Gram matrix in hand, Kernel Ridge Regression is a closed-form
problem:

$$
\boldsymbol{\alpha} \;=\; (K + \lambda I)^{-1}\,\mathbf{y}.
$$

[`quantum_krr_fit()`][quantum_krr_fit] computes this solve and
packages the training covariates, the dual coefficients
$\boldsymbol{\alpha}$ and the kernel configuration into a
`edaphos_quantum_krr` S3 object.

```{r qkrr-fit}
set.seed(1)
train <- sort(sample(length(y), 140L))
test  <- setdiff(seq_along(y), train)

fit <- quantum_krr_fit(
  X[train, ], y[train],
  reps   = 2L,
  lambda = 0.1
)
fit
```

```{r qkrr-predict}
# Numeric scores for soft evaluation
scores <- predict(fit, X[test, ], type = "numeric")
# Hard labels (pm 1) via sign(scores)
pred   <- predict(fit, X[test, ], type = "class")
table(predicted = pred, truth = y[test])
mean(pred == y[test])
```

This is not a benchmark — the point is the **scientific plumbing**:
every piece of the quantum-kernel pipeline is computable in a few
hundred lines of pure R, leaving the mathematical and physical
interpretation visible at every step.

## 6. Interpreting the kernel visually

```{r kernel-heatmap, fig.width=6, fig.height=6}
ord <- order(fit$fitted)
K_sorted <- fit$K_train[ord, ord]

image(seq_len(nrow(K_sorted)), seq_len(ncol(K_sorted)),
      K_sorted[, rev(seq_len(ncol(K_sorted)))],
      xlab = "train sample (ordered by predicted SOC score)",
      ylab = "train sample",
      main = "Quantum kernel — ZZFeatureMap on br_cerrado",
      col  = hcl.colors(64, "YlOrRd"),
      axes = FALSE)
```

Sorting the training samples by their predicted NDVI score reveals
diagonal blocks where the kernel recognises within-class similarity;
off-diagonal entries quantify across-class separation in the quantum
Hilbert space.

## 7. Limits and roadmap

The pure-R simulator scales comfortably to **8 qubits** (256-dim state
vector; fractions of a second per kernel entry). Beyond that,
three avenues open up:

1. **Real NISQ hardware** [@Preskill2018nisq] — the same ZZFeatureMap
   can be shipped as an OpenQASM program to IBM Quantum or similar
   backends. The `edaphos` side contributes the pedometric data
   pipeline and the evaluation metrics.
2. **Variational Quantum Eigensolver (VQE)** [@Peruzzo2014vqe] —
   simulating the Hamiltonian of organo-mineral complexes (clay-humus
   adsorption, iron-oxide weathering) in order to predict soil
   fertility from first principles, rather than by regression from
   surface proxies. This is the direct link to the user's
   "simulation of materials" vision.
3. **Python bridge** (`reticulate` + PennyLane / Qiskit) — when the
   number of qubits exceeds 8, or when differentiable quantum
   parameters are needed (Quantum-PINN, quantum-GPs), the R-side API
   remains the same but the numerical back-end is deferred.

Quantum machine learning in pedometry is unambiguously early. The
role of Pillar 6 in `edaphos` is to provide a clean, self-contained
reference implementation against which more advanced quantum
pedometric workflows can be calibrated.

## References
