---
title: "Pilar 1 — Causal AI: Backdoor Adjustment in Pedogenetic DAGs"
author: "Hugo Rodrigues Machado"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 1 — Causal AI: Backdoor Adjustment in Pedogenetic DAGs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5
)
```

## Abstract

Modern DSM pipelines routinely conflate *variable importance* with
*causal effect*, despite a mature statistical apparatus — Pearl's
structural causal models [@Pearl2009; @Pearl2016primer] — that
distinguishes the two cleanly. The **Pillar 1** of `edaphos` ships a
minimum-viable interface to that apparatus: pedogenetic Directed
Acyclic Graphs (DAGs) expressed in the `dagitty` syntax
[@Textor2016dagitty], automatic backdoor-adjustment set identification,
and a linear causal-effect estimator with a confounded-baseline
comparator. This vignette demonstrates the pipeline on the bundled
Cerrado dataset and highlights the direction of future extensions
toward causal representation learning [@Scholkopf2021] and
time-series causal discovery [@Runge2019].

## 1. Why causal inference in pedometry?

Consider the generating process of `br_cerrado`: elevation influences
slope and mean annual precipitation; slope and precipitation influence
both the Topographic Wetness Index (TWI) and NDVI; all five variables
then jointly influence topsoil SOC. An ordinary regression
\(\mathrm{lm}(\text{soc}\sim\text{ndvi})\) does **not** recover the
direct effect of NDVI on SOC: it sums the direct path with the
backdoor paths through TWI, slope and precipitation
[@Pearl2009, Ch. 3].

Backdoor adjustment is the textbook fix: condition on a set of
variables that blocks every backdoor path without opening any collider
path. Given a DAG, `dagitty::adjustmentSets()` returns valid sets
automatically [@Textor2016dagitty].

## 2. Cerrado-specific DAG

[`causal_cerrado_dag()`][causal_cerrado_dag] encodes the generating
process of `br_cerrado`.

```{r setup, message=FALSE}
library(edaphos)
if (!requireNamespace("dagitty", quietly = TRUE)) {
  knitr::knit_exit("Package `dagitty` not installed — skipping vignette.")
}
g <- causal_cerrado_dag()
g
```

```{r plot-dag}
plot(g)
```

## 3. Adjustment-set identification

```{r conjunto-ajuste}
adj <- causal_adjustment_set(g, exposure = "ndvi", outcome = "soc")
adj
```

The minimal adjustment set (as computed here) blocks every backdoor
path from `ndvi` to `soc` while leaving the direct `ndvi -> soc`
edge unconditioned — a standard back-door criterion result
[@Pearl2009].

## 4. Causal-effect estimation

The wrapper [`causal_estimate_effect()`][causal_estimate_effect] fits
the adjusted linear model
\[
\mathrm{soc}\;=\;\beta_{0}+\beta_{\text{ndvi}}\,\mathrm{ndvi}
                 +\sum_{z\in Z}\gamma_{z}\,z+\varepsilon,
\]
where \(Z\) is the returned adjustment set. It also reports the
confounded-baseline coefficient \(\beta_{\text{ndvi}}^{\text{naive}}\)
obtained from the unadjusted
\(\mathrm{lm}(\text{soc}\sim\text{ndvi})\), so the magnitude of
confounding bias is directly legible.

```{r estimate}
data(br_cerrado, package = "edaphos")
fit <- causal_estimate_effect(
  br_cerrado, g,
  exposure = "ndvi", outcome = "soc",
  effect   = "direct"
)
fit
```

```{r plot-compare, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
df <- data.frame(
  estimator = factor(c("Naive (lm(soc ~ ndvi))",
                        "Adjusted (backdoor)"),
                      levels = c("Naive (lm(soc ~ ndvi))",
                                  "Adjusted (backdoor)")),
  coef      = c(fit$effect_naive, fit$effect)
)
ggplot(df, aes(estimator, coef, fill = estimator)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = signif(coef, 3)), vjust = -0.3) +
  labs(y = "Estimated effect of NDVI on SOC",
       title = "NDVI -> SOC — naive vs. causal (Cerrado DAG)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## 5. CLORPT baseline DAG

For problems beyond `br_cerrado`, [`causal_clorpt_dag()`][causal_clorpt_dag]
returns the classical Jenny (1941) factorial DAG as a template
[@Jenny1941; @McBratney2003]:

```{r clorpt}
g_clorpt <- causal_clorpt_dag()
g_clorpt
```

## 6. How this changes the other pillars

1. **Pillar 5 (Active Learning).** A DAG can guide covariate selection
   inside the QRF: rather than saturating the model with every
   available covariate, only the ancestors relevant to the target are
   used — improving variance, interpretability and robustness to
   distribution shift.
2. **Pillar 2 (PIML).** A pedogenetic DAG restricts which covariates
   can enter the parameterisation of the hierarchical Neural ODE
   [`piml_hierarchical_fit()`][piml_hierarchical_fit], precluding
   spurious-effect learning.
3. **Transfer to new regions.** Where the *causal regime* is shared
   across regions (same ancestors, same edge directions), causally
   adjusted models extrapolate more reliably than purely correlational
   ones [@Scholkopf2021].

## 7. Roadmap

* **Structure learning** via `bnlearn` — learn DAG edges from horizon
  data rather than prescribing them [@Runge2019].
* **Non-linear backdoor estimators** — replace the linear `lm` with
  BART, GAM or a doubly-robust estimator, keeping the same adjustment
  set.
* **Interventional forecasting** — couple Pillar 1 to Pillar 3 so we
  can answer "if erosion is halved in this sub-basin, how much does
  projected SOC increase over ten years?" under explicit causal
  assumptions.

## References
