---
title: "Pilar 5 — Autonomous Active Learning for Soil Mapping"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 5 — Autonomous Active Learning for Soil Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4
)
```

## Abstract

Digital Soil Mapping (DSM) has historically decoupled sampling design
from model fitting: a pre-defined scheme collects observations and a
model is subsequently trained on the resulting dataset
[@McBratney2003; @Minasny2016]. This static pipeline ignores the fact
that, once an initial model exists, its uncertainty field contains
direct guidance on where additional information is most needed
[@Settles2009; @Brus2019]. This vignette formalises the Pillar 5
closed-loop **Active Learning** workflow shipped in `edaphos`: the
algorithm iteratively (i) fits a Quantile Regression Forest
[@Meinshausen2006; @Wright2017ranger], (ii) computes per-candidate
prediction uncertainty and feature-space diversity, (iii) queries the
most informative batch of new locations and (iv) refits after an
*oracle* returns the measured values. We illustrate the pipeline on the
benchmark `meuse` dataset [@Heuvelink2001] and quantify convergence in
Out-of-Bag RMSE under three query strategies plus a logistically
cost-aware variant.

## 1. Statistical framework

Let \(\mathcal{S}\subset\mathbb{R}^{2}\) be the study region and let
\(\mathbf{x}(\mathbf{s})\in\mathbb{R}^{p}\) denote the vector of
environmental covariates at location \(\mathbf{s}\in\mathcal{S}\).
Assume a measurable soil property
\(y(\mathbf{s})=f(\mathbf{x}(\mathbf{s}))+\varepsilon(\mathbf{s})\),
with heteroscedastic error \(\varepsilon\). Given a labelled dataset
\(\mathcal{L}_{t}=\{(\mathbf{s}_{i},\mathbf{x}_{i},y_{i})\}_{i=1}^{n_{t}}\)
and a candidate pool
\(\mathcal{U}_{t}\subset\mathcal{S}\setminus\mathcal{L}_{t}\) at
iteration \(t\), we seek a policy \(\pi\) that selects
\(B_{t}\subset\mathcal{U}_{t}\), \(|B_{t}|=b\), maximising expected
information gain under a finite sampling budget \(N\).

We adopt a Quantile Regression Forest (QRF) [@Meinshausen2006] for
\(\hat{f}_{t}\). For a query probability pair
\((\tau_{\ell},\tau_{u})\in(0,1)^{2}\), the QRF provides a per-pixel
prediction interval
\(\bigl[\hat{q}_{\tau_{\ell}}(\mathbf{x}),\hat{q}_{\tau_{u}}(\mathbf{x})\bigr]\);
the interval width
\[
u_{t}(\mathbf{x})\;=\;\hat{q}_{\tau_{u}}(\mathbf{x})
                    \;-\;\hat{q}_{\tau_{\ell}}(\mathbf{x})
\]
is the Pillar 5 *uncertainty score*.

The *diversity score* is
\[
d_{t}(\mathbf{x})\;=\;\min_{\mathbf{x}^{\prime}\in\mathcal{L}_{t}\cup B_{t}^{<k}}
                      \bigl\|\mathbf{z}(\mathbf{x})-\mathbf{z}(\mathbf{x}^{\prime})\bigr\|_{2},
\]
where \(\mathbf{z}\) denotes a mean-standardised covariate vector and
\(B_{t}^{<k}\) are the candidates already greedily chosen within the
current batch (so the batch is internally diverse).

The **hybrid strategy** combines them through a convex combination of
0-1 min-max-scaled scores:
\[
\pi_{\text{hyb}}(\mathbf{x};\alpha)
   \;=\;\alpha\,\tilde u_{t}(\mathbf{x})
       +(1-\alpha)\,\tilde d_{t}(\mathbf{x}),
\qquad \alpha\in[0,1].
\]
For deployment on autonomous samplers (drones, rovers) we add a
logistical cost term proportional to the Euclidean distance to a base
\(\mathbf{s}_{0}\):
\[
\pi_{\text{cost}}(\mathbf{x};\alpha,\gamma)
   \;=\;\pi_{\text{hyb}}(\mathbf{x};\alpha)
       \;-\;\gamma\,\tilde c(\mathbf{x}),
\qquad c(\mathbf{x})=\|\mathbf{s}(\mathbf{x})-\mathbf{s}_{0}\|_{2}.
\]

These four policies are dispatched by the `strategy` argument of
[`al_query()`][al_query] and [`al_loop()`][al_loop].[^greedy]

[^greedy]: Greedy batch selection is a standard approximation of the
optimal batch policy and is provably near-optimal for submodular
information criteria [@Settles2009].

## 2. Data and seed design

```{r setup, message=FALSE}
library(edaphos)
if (!requireNamespace("sp", quietly = TRUE)) {
  knitr::knit_exit("Package `sp` not installed — skipping vignette.")
}
data(meuse, package = "sp")

d <- stats::na.omit(meuse[, c("x", "y", "dist", "elev", "ffreq",
                              "soil", "lime", "lead")])
d$ffreq <- as.numeric(as.character(d$ffreq))
d$soil  <- as.numeric(as.character(d$soil))
d$lime  <- as.numeric(as.character(d$lime))
```

The `meuse` dataset [@Heuvelink2001] holds \(n = 155\) topsoil samples
from a Dutch floodplain; we use lead concentration (mg kg\(^{-1}\)) as
target and four ancillary covariates. The initial labelled subset is
drawn by **conditioned Latin Hypercube Sampling (cLHS)**
[@Minasny2006clhs], which spreads samples across the joint covariate
distribution and is a strong baseline for DSM [@Brus2019].

```{r seed}
covs <- c("dist", "elev", "ffreq", "soil", "lime")
set.seed(42)
seed_idx <- al_initial_design(d, covariates = covs,
                              n = 15, iter = 1000)
labeled0   <- d[ seed_idx, ]
candidates <- d[-seed_idx, ]
c(n_labeled = nrow(labeled0), n_candidates = nrow(candidates))
```

## 3. Closed-loop fit under the hybrid policy

```{r loop, message=FALSE}
set.seed(42)
model_hybrid <- al_loop(
  labeled    = labeled0,
  candidates = candidates,
  target     = "lead",
  covariates = covs,
  coords     = c("x", "y"),
  budget     = 30, batch = 5,
  strategy   = "hybrid", alpha = 0.7,
  num.trees  = 500L, verbose = FALSE
)
model_hybrid
```

```{r history}
h_hyb <- al_history(model_hybrid)
h_hyb
```

```{r plot-learning, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
ggplot(h_hyb, aes(n_labeled, rmse_oob)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  labs(x = "Labelled sample size n_t",
       y = expression("Out-of-Bag RMSE ("*mg~kg^{-1}*" Pb)"),
       title = "Learning curve under the hybrid policy") +
  theme_minimal(base_size = 12)
```

## 4. Policy comparison

We benchmark the three label-dependent policies while fixing the cLHS
seed and budget (the *ceteris-paribus* principle). Deviation between
curves is attributable to policy choice alone.

```{r compare, message=FALSE}
run_one <- function(strategy) {
  set.seed(42)
  m <- al_loop(
    labeled = labeled0, candidates = candidates,
    target = "lead", covariates = covs, coords = c("x", "y"),
    budget = 30, batch = 5,
    strategy = strategy, alpha = 0.7,
    num.trees = 500L, verbose = FALSE
  )
  cbind(al_history(m), strategy = strategy)
}
h_all <- rbind(
  run_one("uncertainty"),
  run_one("diverse"),
  run_one("hybrid")
)
```

```{r plot-compare, eval=requireNamespace("ggplot2", quietly = TRUE)}
ggplot(h_all, aes(n_labeled, rmse_oob, colour = strategy)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  labs(x = "Labelled sample size n_t",
       y = expression("OOB RMSE ("*mg~kg^{-1}*" Pb)"),
       title = "Query-strategy comparison on meuse (2020 Pb topsoil)",
       colour = "Strategy") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Uncertainty-only sampling exploits the current model maximally but can
cluster redundantly around a single high-variance pocket; pure
diversity explores broadly but wastes budget in regions the model
already handles well [@Settles2009]. The hybrid policy empirically
dominates both in our setting.

## 5. Cost-aware sampling for autonomous platforms

When a drone or rover executes the campaign, the marginal cost of an
observation is *not* uniform. The `"cost"` strategy adds a logistical
penalty proportional to the distance from a base
\(\mathbf{s}_{0}=(x_{0},y_{0})\).

```{r cost, message=FALSE}
set.seed(42)
model_cost <- al_loop(
  labeled = labeled0, candidates = candidates,
  target = "lead", covariates = covs, coords = c("x", "y"),
  budget = 30, batch = 5,
  strategy = "cost",
  base = c(179000, 330000),
  alpha = 0.7, cost_weight = 0.4,
  num.trees = 500L, verbose = FALSE
)
tail(al_history(model_cost), 4)
```

The same decision primitive generalises to on-board *Edge-AI* pipelines
where the oracle is a fielded NIR spectrometer running on the sampler
and the loop closes in seconds rather than weeks.

## 6. Information-theoretic batch acquisition: BatchBALD

The hybrid strategy of §3 scores every candidate by
`alpha * uncertainty + (1 - alpha) * diversity`. It is a well-tuned
heuristic, but a heuristic all the same — there is no formal
objective that it is maximising. **BatchBALD** [@Kirsch2019batchbald]
replaces the heuristic with an information-theoretic objective: pick
the batch that maximises the mutual information between its labels
and the model parameters,

\[
  \mathrm{BatchBALD}(B) \;=\;
  I\bigl(y_B ; \theta \mid x_B, \mathcal D\bigr).
\]

For a regression model with Gaussian aleatoric noise of variance
\(\sigma_a^2\) and an epistemic posterior represented by `T`
parameter draws (for a QRF: the `T` trees of the forest), the
objective reduces to a log-determinant:

\[
  \mathrm{BatchBALD}(B) \;\propto\;
  \tfrac{1}{2}\log\det\!\bigl(
    \mathrm{Cov}_\theta(f_\theta(B)) + \sigma_a^2 I_{|B|}
  \bigr).
\]

The log-det is monotone submodular in \(B\), so a greedy argmax
enjoys the \((1 - 1/e)\) optimality guarantee of
[@Nemhauser1978submodular]. The
[`al_query_batchbald()`][al_query_batchbald] implementation uses
incremental Cholesky / Schur updates so each greedy step is
\(O(m^2 n_{\mathrm{pool}})\) rather than \(O(m^3 n_{\mathrm{pool}})\).

```{r batchbald, eval=TRUE}
library(edaphos)
data(br_cerrado)

covs <- c("elev", "slope", "twi", "map_mm")
idx  <- al_initial_design(br_cerrado, covs, n = 30L, seed = 1L)
lab  <- br_cerrado[idx, ]
pool <- br_cerrado[setdiff(seq_len(nrow(br_cerrado)), idx), ]

fit   <- al_fit(lab, target = "soc", covariates = covs)
batch <- al_query_batchbald(fit, pool, n = 8L)
head(pool[batch, covs])
```

The failure mode that BatchBALD specifically addresses is the
**cluster-of-near-duplicates** pathology of top-\(n\) BALD: when the
pool contains several candidates that are nearly identical in
covariate space, a plain uncertainty-sort returns \(n\) copies of
"the same question." BatchBALD's log-det penalises adding a candidate
whose predictive distribution overlaps strongly with the distribution
of an already-selected candidate, so the greedy step spreads out
through covariate space by construction.

Use BatchBALD when (a) the pool has clustered near-duplicates,
(b) the QRF aleatoric noise is well-estimated (default: out-of-bag
residual variance), and (c) the batch size is moderate (`n <= 50`
for pools of up to ~10 000 candidates). For low-budget hybrid
settings that also need a physical-distance term, the
cost-aware strategy of §3 remains the recommended default.

## 7. Discussion

Active Learning reframes the DSM pipeline from a "collect, then model"
sequence to a feedback controller [@Brus2019]. Three points bear
emphasis:

* **Information-theoretic coherence.** Under Gaussian observation
  noise, maximising the QRF interval width is asymptotically
  equivalent to maximising the mutual information between the next
  observation and the current model — the formal quantity sought by
  Bayesian optimal design [@Settles2009].
* **Coupling with Pillar 2.** Any model-predicted value that falls
  outside a *physically* admissible envelope can be rejected before
  entering the greedy selection (see the `physics_gate` argument of
  [`al_query()`][al_query] and the Pillar 2 vignette); this
  pre-empts pathologies that uncertainty-only strategies otherwise
  amplify.
* **Transferability.** The same abstraction scales from benchmark
  datasets such as `meuse` to planetary-scale products such as
  SoilGrids250m [@HenglWoSIS]; only the oracle implementation changes.

The companion vignette `pilar5-soilgrids-br` demonstrates the identical
workflow on a Cerrado cut-out bundled with `edaphos`.

The hybrid strategy of §3 and the BatchBALD strategy of §6 are
complementary rather than substitutes: the former wins when a
logistical cost term or a physics gate must enter the objective, the
latter wins when the priority is information-theoretically optimal
deduplication across a clustered pool. Both share the
`physics_gate` interface, so any Pillar 2 PIML rejection rule plugs
into both without change.

## References
