---
title: "Pilar 4 — Contrastive Raster Embeddings as Active-Learning Covariates"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 4 — Contrastive Raster Embeddings as Active-Learning Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4.2
)
```

## Abstract

Self-Supervised Learning (SSL) has emerged as the dominant
representation-learning paradigm in computer vision
[@Chen2020simclr; @He2020moco; @Oord2018cpc] and is increasingly
applied to remote sensing [@Jean2019tile2vec; @ReichsteinDL2019]. The
**Pillar 4** of `edaphos` ships a minimal SimCLR-style pipeline
[@Chen2020simclr] that pre-trains a small raster encoder on unlabelled
covariate patches; the resulting per-pixel embeddings are then used as
additional covariates inside the Pillar 5 Active Learning loop on
`br_cerrado`. We report a head-to-head comparison of the identical AL
policy with and without the learned features.

## 1. Contrastive representation learning

Given a mini-batch \(\{\mathbf{x}_{i}\}_{i=1}^{B}\) of raster patches,
SimCLR draws two independent stochastic augmentations
\(\tilde{\mathbf{x}}_{i}^{(1)},\tilde{\mathbf{x}}_{i}^{(2)}\) of each
patch, encodes them with a shared CNN \(f_{\theta}\) and a projection
head \(g_{\phi}\), and minimises the normalised temperature-scaled
cross-entropy (NT-Xent) loss [@Chen2020simclr]:
\[
\mathcal{L}_{\text{NT-Xent}}
   \;=\;-\frac{1}{2B}
         \sum_{i=1}^{B}\sum_{v\in\{1,2\}}
           \log\frac{\exp(\mathrm{sim}(\mathbf{z}_{i}^{(v)},
                                         \mathbf{z}_{i}^{(v^{\prime})})/\tau)}
           {\sum_{k\ne(i,v)}\exp(\mathrm{sim}(\mathbf{z}_{i}^{(v)},
                                              \mathbf{z}_{k})/\tau)},
\]
where \(\mathbf{z}=g_{\phi}(f_{\theta}(\tilde{\mathbf{x}}))\),
\(\mathrm{sim}(\cdot,\cdot)\) is cosine similarity and \(\tau\) is the
temperature. After pre-training, the projection head is discarded and
the *backbone* feature vector \(f_{\theta}(\mathbf{x})\) is used as the
reusable representation.

## 2. Data preparation

Each of the 2025 pixels of `br_cerrado` is made the centre of a
\(7\times 7\times 5\) patch with five normalised channels (elevation,
slope, TWI, mean annual precipitation, NDVI). Reflection padding
preserves the original grid size.

```{r setup, message=FALSE}
library(edaphos)
.torch_ok <- requireNamespace("torch", quietly = TRUE) &&
             isTRUE(tryCatch(torch::torch_is_installed(),
                             error = function(e) FALSE))
if (!.torch_ok) {
  knitr::knit_exit(
    "torch runtime (libtorch) not available — skipping vignette."
  )
}
data(br_cerrado, package = "edaphos")

covs_base <- c("elev", "slope", "twi", "map_mm", "ndvi")
H <- length(unique(br_cerrado$y))
W <- length(unique(br_cerrado$x))
stopifnot(H * W == nrow(br_cerrado))
c(H = H, W = W)
```

```{r patchify}
grid <- br_cerrado[order(br_cerrado$y, br_cerrado$x), ]
mat <- function(col) matrix(grid[[col]], nrow = H, ncol = W, byrow = TRUE)
z_mat <- function(m) (m - mean(m)) / max(stats::sd(m), 1e-6)
channels <- lapply(covs_base, function(cc) z_mat(mat(cc)))
cube <- array(unlist(channels), dim = c(H, W, length(covs_base)))
cube <- aperm(cube, c(3L, 1L, 2L))
mirror_pad <- function(m, p) {
  top <- m[p:1, , drop = FALSE]
  bot <- m[nrow(m):(nrow(m) - p + 1L), , drop = FALSE]
  tall <- rbind(top, m, bot)
  left  <- tall[, p:1, drop = FALSE]
  right <- tall[, ncol(tall):(ncol(tall) - p + 1L), drop = FALSE]
  cbind(left, tall, right)
}
patch_size <- 7L
pad <- (patch_size - 1L) %/% 2L
C <- dim(cube)[1L]
padded <- array(0, dim = c(C, H + 2L * pad, W + 2L * pad))
for (cc in seq_len(C)) padded[cc, , ] <- mirror_pad(cube[cc, , ], pad)
patches <- array(0, dim = c(H * W, C, patch_size, patch_size))
k <- 1L
for (i in seq_len(H)) for (j in seq_len(W)) {
  patches[k, , , ] <- padded[, i:(i + 2L * pad), j:(j + 2L * pad)]
  k <- k + 1L
}
dim(patches)
```

## 3. SimCLR pre-training

Four hundred randomly-drawn patches are pre-trained for 60 epochs on
Adam. The backbone is a two-layer CNN (32 feature channels) followed
by global average pooling.

```{r pretrain, message=FALSE}
set.seed(1)
pretrain_idx <- sample(nrow(patches), 400L)
simclr <- foundation_simclr_pretrain(
  patches[pretrain_idx, , , , drop = FALSE],
  feature_dim = 24L, proj_dim  = 12L,
  batch_size  = 32L, epochs    = 60L, lr = 0.005,
  temperature = 0.2, noise_sd  = 0.1,
  seed        = 1L, verbose    = FALSE
)
simclr
```

## 4. Per-pixel embedding extraction

```{r embed}
emb <- foundation_simclr_embed(simclr, patches)
dim(emb)
colnames(emb) <- paste0("emb", sprintf("%02d", seq_len(ncol(emb))))
br_ext <- cbind(grid, emb)
head(br_ext[, c(covs_base, "soc", head(colnames(emb), 3L))])
```

## 5. Controlled experiment — AL with and without SimCLR features

Same cLHS seed, same budget, same hybrid policy; only the covariate
list changes. Any difference in learning curves is attributable to the
learned features.

```{r al-compare, message=FALSE}
set.seed(42)
seed_idx <- al_initial_design(br_ext, covariates = covs_base,
                               n = 25L, iter = 1500L)
run_al <- function(covariate_set, label) {
  set.seed(42)
  m <- al_loop(
    labeled    = br_ext[ seed_idx, ],
    candidates = br_ext[-seed_idx, ],
    target     = "soc", covariates = covariate_set,
    coords     = c("x", "y"),
    budget     = 45L, batch = 5L,
    strategy   = "hybrid", alpha = 0.7,
    num.trees  = 500L, verbose = FALSE
  )
  cbind(al_history(m), setup = label)
}
covs_ext <- c(covs_base, colnames(emb))
h_base <- run_al(covs_base, "baseline (5 covariates)")
h_ext  <- run_al(covs_ext,  "baseline + SimCLR (29 covariates)")
h <- rbind(h_base, h_ext)
tail(h[h$setup == "baseline (5 covariates)", ], 4)
tail(h[h$setup == "baseline + SimCLR (29 covariates)", ], 4)
```

```{r plot-compare, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
ggplot(h, aes(n_labeled, rmse_oob, colour = setup)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  labs(x = "Labelled sample size n_t",
       y = expression("OOB RMSE ("*g~kg^{-1}*" SOC)"),
       title = "AL on br_cerrado — SimCLR features as auxiliary covariates",
       colour = "Feature set") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 6. Discussion

On this small synthetic grid (N = 2025 pixels, five physically-coded
covariates, 60 SimCLR epochs) the learned features typically yield a
marginal OOB-RMSE improvement, which is itself informative: the
encoder extracts signal from the covariate stack **with zero labels
consumed during pre-training** [@Jean2019tile2vec].

The payoff scales super-linearly with the size of the unlabelled
corpus. Under a realistic Cerrado-wide pre-training with optical + SAR
+ DEM + climate inputs, the learned representation captures transfer
structure unavailable to any regional model [@Chen2020simclr; @ReichsteinDL2019],
and becomes the de-facto feature layer for Pillars 2, 3 and 5. That
planetary-scale workflow is scheduled for the next major release.

## References
