---
title: "Pilar 1 — Causal AI: Backdoor Adjustment in Pedogenetic DAGs"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 1 — Causal AI: Backdoor Adjustment in Pedogenetic DAGs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5
)
```

## Abstract

Modern DSM pipelines routinely conflate *variable importance* with
*causal effect*, despite a mature statistical apparatus — Pearl's
structural causal models [@Pearl2009; @Pearl2016primer] — that
distinguishes the two cleanly. The **Pillar 1** of `edaphos` ships a
minimum-viable interface to that apparatus: pedogenetic Directed
Acyclic Graphs (DAGs) expressed in the `dagitty` syntax
[@Textor2016dagitty], automatic backdoor-adjustment set identification,
and a linear causal-effect estimator with a confounded-baseline
comparator. This vignette demonstrates the pipeline on the bundled
Cerrado dataset and highlights the direction of future extensions
toward causal representation learning [@Scholkopf2021] and
time-series causal discovery [@Runge2019].

## 1. Why causal inference in pedometry?

Consider the generating process of `br_cerrado`: elevation influences
slope and mean annual precipitation; slope and precipitation influence
both the Topographic Wetness Index (TWI) and NDVI; all five variables
then jointly influence topsoil SOC. An ordinary regression
\(\mathrm{lm}(\text{soc}\sim\text{ndvi})\) does **not** recover the
direct effect of NDVI on SOC: it sums the direct path with the
backdoor paths through TWI, slope and precipitation
[@Pearl2009, Ch. 3].

Backdoor adjustment is the textbook fix: condition on a set of
variables that blocks every backdoor path without opening any collider
path. Given a DAG, `dagitty::adjustmentSets()` returns valid sets
automatically [@Textor2016dagitty].

## 2. Cerrado-specific DAG

[`causal_cerrado_dag()`][causal_cerrado_dag] encodes the generating
process of `br_cerrado`.

```{r setup, message=FALSE}
library(edaphos)
if (!requireNamespace("dagitty", quietly = TRUE)) {
  knitr::knit_exit("Package `dagitty` not installed — skipping vignette.")
}
g <- causal_cerrado_dag()
g
```

```{r plot-dag}
plot(g)
```

## 3. Adjustment-set identification

```{r conjunto-ajuste}
adj <- causal_adjustment_set(g, exposure = "ndvi", outcome = "soc")
adj
```

The minimal adjustment set (as computed here) blocks every backdoor
path from `ndvi` to `soc` while leaving the direct `ndvi -> soc`
edge unconditioned — a standard back-door criterion result
[@Pearl2009].

## 4. Causal-effect estimation

The wrapper [`causal_estimate_effect()`][causal_estimate_effect] fits
the adjusted linear model
\[
\mathrm{soc}\;=\;\beta_{0}+\beta_{\text{ndvi}}\,\mathrm{ndvi}
                 +\sum_{z\in Z}\gamma_{z}\,z+\varepsilon,
\]
where \(Z\) is the returned adjustment set. It also reports the
confounded-baseline coefficient \(\beta_{\text{ndvi}}^{\text{naive}}\)
obtained from the unadjusted
\(\mathrm{lm}(\text{soc}\sim\text{ndvi})\), so the magnitude of
confounding bias is directly legible.

```{r estimate}
data(br_cerrado, package = "edaphos")
fit <- causal_estimate_effect(
  br_cerrado, g,
  exposure = "ndvi", outcome = "soc",
  effect   = "direct"
)
fit
```

```{r plot-compare, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
df <- data.frame(
  estimator = factor(c("Naive (lm(soc ~ ndvi))",
                        "Adjusted (backdoor)"),
                      levels = c("Naive (lm(soc ~ ndvi))",
                                  "Adjusted (backdoor)")),
  coef      = c(fit$effect_naive, fit$effect)
)
ggplot(df, aes(estimator, coef, fill = estimator)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = signif(coef, 3)), vjust = -0.3) +
  labs(y = "Estimated effect of NDVI on SOC",
       title = "NDVI -> SOC — naive vs. causal (Cerrado DAG)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## 5. CLORPT baseline DAG

For problems beyond `br_cerrado`, [`causal_clorpt_dag()`][causal_clorpt_dag]
returns the classical Jenny (1941) factorial DAG as a template
[@Jenny1941; @McBratney2003]:

```{r clorpt}
g_clorpt <- causal_clorpt_dag()
g_clorpt
```

## 6. LLM-driven Knowledge-Graph augmentation

The expert-written DAG above encodes a single analyst's view of
pedogenesis. A more defensible structural model would reflect the
wider published literature — but manually encoding every causal claim
from hundreds of papers is infeasible. Pillar 1 therefore ships a
three-stage **LLM + Knowledge-Graph (KG) pipeline** that automates the
transition from a corpus of abstracts to an augmented DAG:

```
  abstracts  ->  causal_llm_extract()  ->  tidy claims  ->  causal_kg_*  ->  causal_augment_dag()
  (corpus)     (Gemma-4 / GPT-5 / Claude)   (data frame)    (KG)          (augmented DAG)
```

Three interchangeable backends are supported behind a single
extractor:

* `"ollama"` — local, zero-cost. Default model `"gemma4:latest"`
  (9.6 GB, e4b); switch to `"gemma4:26b"` for higher extraction
  fidelity. This is the backend used to produce the snapshot shipped
  in `inst/extdata/cerrado_claims.jsonl`.
* `"openai"` — hosted GPT family with JSON mode. Requires
  `OPENAI_API_KEY`.
* `"anthropic"` — Claude Messages API. Requires `ANTHROPIC_API_KEY`.

### 6.1 Corpus and cached extractions

Ten curated Cerrado-pedology abstracts ship in
`inst/extdata/cerrado_abstracts.jsonl`, together with the Gemma-4
extractions ready to load from `inst/extdata/cerrado_claims.jsonl`.
Loading the cached JSON lets the vignette build offline, reproducibly,
even on CI runners that do not have Ollama.

```{r load-corpus, message=FALSE}
if (!requireNamespace("igraph", quietly = TRUE)) {
  knitr::knit_exit("Package `igraph` not installed — skipping KG section.")
}
abs_path <- system.file("extdata", "cerrado_abstracts.jsonl",
                        package = "edaphos")
claim_path <- system.file("extdata", "cerrado_claims.jsonl",
                          package = "edaphos")

abstracts <- do.call(rbind, lapply(readLines(abs_path), function(l) {
  j <- jsonlite::fromJSON(l)
  data.frame(source = j$source, abstract = j$abstract,
             stringsAsFactors = FALSE)
}))
claims <- do.call(rbind, lapply(readLines(claim_path), function(l) {
  as.data.frame(jsonlite::fromJSON(l), stringsAsFactors = FALSE)
}))
nrow(abstracts); nrow(claims)
head(claims[, c("cause", "effect", "source", "confidence")], 5)
```

### 6.2 Re-running the live extractor (optional)

When an Ollama server is available at `localhost:11434`, the same
claims are recovered end-to-end. Guard the chunk so the vignette is
still reproducible on a machine without Ollama:

```{r live-extract, eval=FALSE}
kg_live <- causal_kg_new()
kg_live <- causal_llm_ingest_corpus(
  kg_live, abstracts,
  abstract_col = "abstract",
  source_col   = "source",
  backend      = "ollama",
  model        = "gemma4:latest",
  min_confidence = 0.5,
  progress = function(i, n, s) message(sprintf("[%d/%d] %s", i, n, s))
)
kg_live
```

### 6.3 Building the KG from the cached claims

```{r build-kg}
kg <- causal_kg_new()
for (i in seq_len(nrow(claims))) {
  kg <- causal_kg_add_edge(
    kg,
    cause      = claims$cause[i],
    effect     = claims$effect[i],
    source     = claims$source[i],
    evidence   = claims$evidence[i],
    confidence = claims$confidence[i]
  )
}
kg
```

### 6.4 Augmenting the Cerrado DAG

`causal_augment_dag()` takes the expert base DAG and unions it with
every KG edge whose confidence is `>= min_confidence`. Edges that
would introduce a cycle — which would break backdoor identifiability
— are rejected with a warning.

```{r augment-dag}
g_aug <- causal_augment_dag(g, kg, min_confidence = 0.7)
diff <- causal_augment_diff(g, g_aug)
table(diff$origin)
head(diff[diff$origin == "kg", ], 8)
```

### 6.5 Re-estimating the effect with the augmented DAG

If the literature agrees that a previously-ignored variable
(`erosion`, say) confounds `ndvi -> soc`, the augmented DAG will add
it to the minimal adjustment set, and the point estimate of the
direct effect may shift materially.

```{r augmented-effect}
adj_base <- causal_adjustment_set(g,     exposure = "ndvi", outcome = "soc")
adj_aug  <- tryCatch(
  causal_adjustment_set(g_aug, exposure = "ndvi", outcome = "soc"),
  error = function(e) adj_base
)
list(base = adj_base, augmented = adj_aug)
```

The remainder of the analysis — fitting an adjusted model, comparing
against a naive estimator — is unchanged: the DAG is just richer.

## 6.6 Ontology alignment

LLM output labels are free-form — one abstract produces
`steeper_slopes`, another `slope_angle`, a third `slope`. Before a
Knowledge Graph fuses with a classical DAG, those synonymous labels
must collapse to a single canonical term. Pillar 1 ships a three-tier
matcher — exact, substring, fuzzy Levenshtein — against a hand-curated
Cerrado-pedometry vocabulary (~60 terms, subset of AGROVOC + ENVO):

```{r align, eval=requireNamespace("igraph", quietly = TRUE)}
mapping <- causal_kg_alignment(kg)
head(mapping[mapping$method != "exact", ], 8)
kg_canon <- causal_kg_rename(kg, mapping)
kg_canon
```

Live SPARQL queries to the FAO AGROVOC endpoint
(`causal_ontology_agrovoc()`) and parsing of a local ENVO `.obo` file
(`causal_ontology_envo()`, via the optional `ontologyIndex`
Suggests) are exposed for problems whose vocabulary extends beyond
the Cerrado core.

## 7. Corpus ingestion from SciELO and OpenAlex

Two zero-account clients turn literature search queries into
abstract-ready data frames that can be piped directly into
[`causal_llm_ingest_corpus()`][causal_llm_ingest_corpus]:

```{r corpus, eval=FALSE}
# SciELO — Latin-American scientific literature (keyless)
scielo_df <- causal_corpus_scielo(
  query        = "Cerrado soil organic carbon",
  max_results  = 30L,
  from_year    = 2015
)

# OpenAlex — global scholarly graph (keyless; a `mailto=` tier raises the
# rate limit). Abstracts are reconstructed from OpenAlex's inverted index.
openalex_df <- causal_corpus_openalex(
  query       = "Cerrado soil organic carbon",
  max_results = 30L,
  mailto      = "you@example.org"
)

# Feed straight into the LLM pipeline:
kg_live <- causal_llm_ingest_corpus(
  causal_kg_new(),
  rbind(scielo_df, openalex_df),
  abstract_col = "abstract", source_col = "source",
  backend = "ollama", model = "gemma4:latest"
)
```

Both clients return identical column schemas (`source`, `title`,
`abstract`, `year`, `doi`, `url`) so downstream code is agnostic to
the upstream database.

## 8. Non-linear effect estimation via BART

The default `estimator = "lm"` works when the covariate–outcome
relationship is approximately linear once the adjustment set is
conditioned on. For non-linear regimes the same function accepts
`estimator = "bart"` and fits Bayesian Additive Regression Trees
(Chipman, George and McCulloch 2010) on the same adjustment set. The
direct effect is computed as the average partial derivative

\[
\bar\partial = \frac{1}{n}\sum_{i=1}^{n}
  \frac{\widehat{E}[Y \mid X = x_i + \delta,\, Z = z_i]
        - \widehat{E}[Y \mid X = x_i,\, Z = z_i]}{\delta},
\]

averaged over the training data; a 95 % credible interval is
recovered from the BART posterior.

```{r bart, eval=requireNamespace("dbarts", quietly = TRUE)}
set.seed(1)
fit_bart <- causal_estimate_effect(
  br_cerrado, g,
  exposure = "ndvi", outcome = "soc",
  estimator = "bart",
  bart_kwargs = list(ndpost = 200L, nskip = 100L)
)
fit_bart
```

On the linear-by-construction `br_cerrado` DGP the BART point
estimate agrees with the OLS backdoor estimate within the posterior
credible width — a reassuring sanity check that the non-linear
estimator does not over-fit the adjustment set.

## 9. How this changes the other pillars

1. **Pillar 5 (Active Learning).** A DAG can guide covariate selection
   inside the QRF: rather than saturating the model with every
   available covariate, only the ancestors relevant to the target are
   used — improving variance, interpretability and robustness to
   distribution shift.
2. **Pillar 2 (PIML).** A pedogenetic DAG restricts which covariates
   can enter the parameterisation of the hierarchical Neural ODE
   [`piml_hierarchical_fit()`][piml_hierarchical_fit], precluding
   spurious-effect learning.
3. **Transfer to new regions.** Where the *causal regime* is shared
   across regions (same ancestors, same edge directions), causally
   adjusted models extrapolate more reliably than purely correlational
   ones [@Scholkopf2021].

## 10. Scaling to tens of thousands of abstracts

Section 6 fed the LLM extractor a curated set of ten Cerrado-pedology
abstracts. Production usage -- e.g. building a Cerrado-wide Knowledge
Graph from the bulk of the last twenty years of published soil-science
literature -- pushes the corpus size three to four orders of magnitude
higher. Pillar 1 therefore ships three production-grade helpers:

### 10.1 Paginated corpus clients

Both [`causal_corpus_openalex()`][causal_corpus_openalex] and
[`causal_corpus_scielo()`][causal_corpus_scielo] now page transparently
through the upstream APIs (cursor-based for OpenAlex, limit-offset for
SciELO) until `max_results` is reached. A multi-thousand pull is a
single R call:

```{r scale-query, eval=FALSE}
corpus <- rbind(
  causal_corpus_openalex("Brazilian Cerrado soil organic carbon",
                          max_results = 5000L,
                          mailto      = "you@example.org"),
  causal_corpus_openalex("Cerrado land-use change soil fertility",
                          max_results = 5000L,
                          mailto      = "you@example.org")
)
corpus <- causal_corpus_deduplicate(corpus)  # DOI / title dedup
nrow(corpus)
```

### 10.2 Resumable, cached LLM ingestion

At 10 000 abstracts and five seconds per LLM call, an unbroken
ingestion is a half-day run. The `cache_dir` and `max_retries`
arguments of [`causal_llm_ingest_corpus()`][causal_llm_ingest_corpus]
make that operationally possible: every `(source, abstract)` pair is
hashed to a stable filename, and cached JSON responses short-circuit
the LLM call on re-runs. Failed calls (malformed JSON, timeouts) are
retried up to `max_retries` times with exponential backoff; rows that
exhaust all retries are reported in `attr(kg, "failed")`.

```{r scale-ingest, eval=FALSE}
kg <- causal_kg_new()
kg <- causal_llm_ingest_corpus(
  kg, corpus,
  backend        = "ollama",
  model          = "gemma4:latest",
  min_confidence = 0.6,
  cache_dir      = "~/edaphos-llm-cache",  # survives crashes
  max_retries    = 2L
)
length(attr(kg, "failed"))
```

A bundled 100-abstract Cerrado run ships with the package as
`inst/extdata/cerrado_claims_real_corpus.jsonl` so users can inspect
the end product without re-running the LLM themselves.

### 10.3 Live AGROVOC alignment

Once the KG has grown to several hundred nodes, the curated Cerrado
vocabulary of [`causal_ontology_cerrado()`][causal_ontology_cerrado]
stops covering the tail. `causal_kg_alignment()` therefore accepts
`vocab = "agrovoc"`, which hits the FAO SPARQL endpoint live for
every unique node, picks the Levenshtein-nearest `skos:prefLabel`,
and caches the result on disk:

```{r agrovoc-align, eval=FALSE}
mapping <- causal_kg_alignment(
  kg,
  vocab         = "agrovoc",
  agrovoc_cache = "~/edaphos-agrovoc-cache.rds"
)
head(mapping[!is.na(mapping$uri), c("original", "canonical", "uri")])

kg_canonical <- causal_kg_rename(kg, mapping)
```

The returned data frame has an extra `uri` column pointing at the
AGROVOC concept URI -- `http://aims.fao.org/aos/agrovoc/c_xxxxx` --
so the resulting KG is directly inter-operable with other AGROVOC-
aligned pipelines (FAOSTAT, Agrisemantics, the Semantic Agronomy
project).

## 11. Paper-scale audit: persistence, Turtle, ranked edges

Ingesting tens of thousands of abstracts is only useful if the
resulting Knowledge Graph survives long enough to be interrogated.
As of **v1.0.0** `edaphos` ships three primitives that turn a
large KG from an opaque in-memory object into a paper-scale research
artefact: **persistence**, **RDF export**, and **multi-source
ranking**.

### 11.1 Persistence

`causal_kg_save()` / `causal_kg_load()` serialise a KG through its
tidy edge list — not through `igraph`'s raw C-level pointer layout —
so the resulting `.rds` is portable across `igraph` versions and
byte-reproducible:

```{r kg-save-roundtrip, eval=requireNamespace("igraph", quietly = TRUE)}
kg_roundtrip <- causal_kg_new()
kg_roundtrip <- causal_kg_add_edge(kg_roundtrip, "precipitation", "soc",
                                     source = "Jenny 1941", confidence = 0.9)
kg_roundtrip <- causal_kg_add_edge(kg_roundtrip, "precipitation", "soc",
                                     source = "Minasny 2017", confidence = 0.85)
kg_roundtrip <- causal_kg_add_edge(kg_roundtrip, "slope", "soc",
                                     source = "Jenny 1941", confidence = 0.70)

f <- tempfile(fileext = ".rds")
causal_kg_save(kg_roundtrip, f)
kg_loaded <- causal_kg_load(f)
identical(causal_kg_edges(kg_roundtrip), causal_kg_edges(kg_loaded))
```

### 11.2 RDF 1.1 Turtle export

`causal_kg_to_turtle()` emits a self-contained, W3C-conformant
RDF 1.1 Turtle document [@Beckett2014turtle] with a reified
`rdf:Statement` per causal edge. All provenance — confidence,
evidence, source(s), timestamp — is preserved in a SPARQL-queryable
form, and each node gets a stable IRI inside a user-controlled
namespace so the KG can be federated with external vocabularies
(AGROVOC, ENVO, SKOS thesauri) by a simple `owl:sameAs`:

```{r kg-turtle, eval=requireNamespace("igraph", quietly = TRUE)}
ttl <- causal_kg_to_turtle(
  kg_roundtrip,
  base_uri   = "https://edaphos.io/kg/",
  namespaces = c(agrovoc = "http://aims.fao.org/aos/agrovoc/")
)
cat(substr(ttl, 1, 600))
```

The emitter is pure R (no RDF library dependency). The output is
guaranteed parseable by any RDF 1.1-conformant consumer: rdflib,
Apache Jena, Oxigraph, Blazegraph, GraphDB, Virtuoso.

### 11.3 Multi-source edge ranking

On a KG built from thousands of papers the critical audit question is
**which causal claims are supported by the most independent
sources** — an edge asserted by 50 papers with mean confidence 0.8 is
far more trustworthy than one asserted by a single paper with
confidence 1.0. `causal_kg_rank_edges()` collapses the KG to unique
`(cause, effect)` pairs and sorts by a priority list of metrics:

```{r kg-rank, eval=requireNamespace("igraph", quietly = TRUE)}
causal_kg_rank_edges(
  kg_roundtrip,
  by = c("n_sources", "mean_confidence")
)
```

When an alignment data frame is supplied, an `agrovoc_support`
column is attached (0 / 0.5 / 1 depending on how many of the edge's
endpoints resolve to an AGROVOC concept) so the caller can prefer
edges whose nodes are anchored in a community-governed vocabulary.

`summary.edaphos_causal_kg()` is the one-line health check — node
and edge counts, number of unique sources, confidence quartiles and
a DAG-ness verdict — useful right after a 10 k-abstract
`causal_llm_ingest_corpus()` run:

```{r kg-summary, eval=requireNamespace("igraph", quietly = TRUE)}
summary(kg_roundtrip)
```

### 11.4 Concurrent AGROVOC alignment

Aligning a 10 k-node KG one term at a time against the FAO AGROVOC
SPARQL endpoint takes hours. v1.0.0 introduces
`causal_ontology_agrovoc_align_batch()`, which batches at the
**transport layer** via `httr2::req_perform_parallel()` — we chose
this approach because AGROVOC's production endpoint rejects genuine
SPARQL-level batching (`VALUES` + `CONTAINS(?label, ?term)`) with a
504 gateway timeout, since the substring filter cannot short-circuit
against a bound term set.

```{r kg-agrovoc-batch, eval=FALSE}
vocab <- unique(c(causal_kg_edges(kg_big)$cause,
                  causal_kg_edges(kg_big)$effect))
ag <- causal_ontology_agrovoc_align_batch(
  vocab,
  cache_path = "tools/.agrovoc_cache.rds",
  max_active = 5L,       # number of concurrent HTTP connections
  max_retries = 2L,
  verbose = TRUE
)
mean(!is.na(ag$uri))      # AGROVOC resolution rate
```

The on-disk cache is idempotent: a second call over the same
vocabulary replays every match from disk with ~0 network traffic.
`causal_kg_alignment(kg, vocab = "agrovoc", agrovoc_batch = TRUE)`
is the KG-level one-liner that wires the batched resolver through
the standard alignment output.

## 12. Roadmap

* **Structure learning** via `bnlearn` — learn DAG edges from horizon
  data rather than prescribing them [@Runge2019].
* **Non-linear backdoor estimators** — replace the linear `lm` with
  BART, GAM or a doubly-robust estimator, keeping the same adjustment
  set.
* **Interventional forecasting** — couple Pillar 1 to Pillar 3 so we
  can answer "if erosion is halved in this sub-basin, how much does
  projected SOC increase over ten years?" under explicit causal
  assumptions.
* **Conflict-resolved ingestion** — majority voting across multiple
  LLM extractors (Ollama / OpenAI / Anthropic) or Bayesian
  confidence pooling to arbitrate disagreeing causal claims over
  the same abstract.

## References
