---
title: "Pilar 1 on real data — identified causal effects on Cerrado SOC"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 1 on real data — identified causal effects on Cerrado SOC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7.5, fig.height = 4.5,
  fig.align = "center", out.width = "92%", dpi = 150
)
```

## 1. Why this vignette exists

The original Pillar 1 vignette (`vignette("pilar1-causal")`) derives
the backdoor-adjustment machinery from Pearl 2009 and demonstrates it
on the `br_cerrado` synthetic dataset that ships with the package.
As of **v1.4.0** the Pillar 1 story runs on **real Brazilian
Cerrado data**: 1 095 WoSIS topsoil profiles
[@Batjes2020wosis], the same dataset the v1.3.1 case-study benchmark
consumes. The three quantities below are publishable causal
estimates, not pedagogic toys.

```{r load, message = FALSE}
library(edaphos)
library(ggplot2)
library(dplyr)

res_path <- system.file("extdata", "causal_cerrado_real.rds",
                         package = "edaphos")
stopifnot(nzchar(res_path), file.exists(res_path))
R <- readRDS(res_path)
```

## 2. The DAG used for identification

[`causal_cerrado_real_dag()`](../reference/causal_cerrado_real_dag.html)
encodes published Cerrado pedogenetic knowledge on the exact column
names of the v1.3.1 case-study bundle. It is a DAG over 12 nodes
covering relief, climate, texture, bulk density and land-cover
fractions, with 23 directed edges.

```{r dag, fig.cap = "Cerrado real-data DAG. Arrows encode six classes of soil-formation relations: relief -> climate (orographic), climate -> vegetation, relief -> texture (erosion), texture -> density, climate + texture + slope + land cover -> topsoil SOC.", fig.height = 5.6}
if (requireNamespace("ggdag", quietly = TRUE)) {
  ggdag::ggdag(R$dag, layout = "sugiyama", text_size = 2.7) +
    ggdag::theme_dag_blank()
} else {
  # `plot(dagitty)` requires node coordinates, which a bare DAG does
  # not carry. Fall back to a textual list of the edges.
  cat("Install the `ggdag` package to see the graph.\n\n",
      paste(dagitty::edges(R$dag)$v, "->",
             dagitty::edges(R$dag)$w, collapse = "\n"))
}
```

## 3. Three exposures of practical interest

For each exposure we:

1. Compute the minimal backdoor-adjustment set from the DAG
   via [`causal_adjustment_set()`](../reference/causal_adjustment_set.html);
2. Fit a linear backdoor-adjusted direct-effect estimator via
   [`causal_estimate_effect(estimator = "lm")`](../reference/causal_estimate_effect.html);
3. Repeat with a non-linear BART estimator [@Chipman2010bart]
   (`estimator = "bart"`);
4. Compute a **block bootstrap** confidence interval resampling
   k-means clusters of the 5-fold CV assignment — the honest
   alternative to the asymptotic `confint(lm)` interval when the
   data are spatially clustered.

### 3.1 Mean annual precipitation → SOC (climate forcing)

Exposure: **`wc_bio_12`** (mean annual precipitation, mm/year).
Outcome: `soc_topsoil_gkg`.

```{r show-bio12}
e <- R$effects$wc_bio_12
cat("Adjustment set:", paste(e$adjustment, collapse = ", "), "\n")
cat(sprintf("  LM  naive = %+.4f g/kg per mm\n", e$lm$effect_naive))
cat(sprintf("  LM  direct = %+.4f  CI95 = [%+.4f, %+.4f]  boot = [%+.4f, %+.4f]\n",
            e$lm$effect, e$lm$effect_ci[1L], e$lm$effect_ci[2L],
            e$lm$effect_boot_ci[1L], e$lm$effect_boot_ci[2L]))
cat(sprintf("  BART direct (at +IQR/2) = %+.4f  CI95 = [%+.4f, %+.4f]\n",
            e$bart$effect, e$bart$effect_ci[1L], e$bart$effect_ci[2L]))
```

**Interpretation.** At the canonical Cerrado MAP gradient (≈ 500 mm
range between Cerrado-Caatinga transition and Cerrado-Amazon
transition), the **identified direct effect** predicts an increase
of roughly **`r round(500 * e$lm$effect, 1)` g/kg SOC** across that
gradient. The naive OLS slope and the identified slope coincide to
4 decimal places in this case — the backdoor confounders
(temperature, slope, land cover) happen to be near-orthogonal to
precipitation within the Cerrado AoI.

### 3.2 Tree-cover fraction → SOC (land-use pathway)

Exposure: **`wc_landcover_trees`** (% tree cover from ESA WorldCover
2020, 0–100 % fractional).

```{r show-trees}
e <- R$effects$wc_landcover_trees
cat("Adjustment set:", paste(e$adjustment, collapse = ", "), "\n")
cat(sprintf("  LM  naive = %+.4f g/kg per %%   -> direct = %+.4f  boot = [%+.4f, %+.4f]\n",
            e$lm$effect_naive, e$lm$effect,
            e$lm$effect_boot_ci[1L], e$lm$effect_boot_ci[2L]))
cat(sprintf("  BART direct (at +IQR/2) = %+.4f  CI95 = [%+.4f, %+.4f]\n",
            e$bart$effect, e$bart$effect_ci[1L], e$bart$effect_ci[2L]))
```

**Interpretation.** The naive OLS says tree cover lifts topsoil SOC
by +0.9 g/kg per 1 % tree cover. After backdoor-adjusting for
climate (bio_01 and bio_12, which jointly determine the land-cover
fraction as well as SOC directly), **the identified direct effect
is +2.05 g/kg per 1 % tree cover** — *more than double* the naive
slope. The confounder structure here pulls the association
*downwards*: warmer / drier Cerrado cells have less tree cover AND
less SOC, making the raw correlation look weaker than the true
causal effect. The bootstrap CI is wide (crosses zero at its lower
edge) because the subsets of the biome with very high tree cover
are clustered.

### 3.3 Clay % → SOC (mineral-protection mechanism)

Exposure: **`soilgrids_clay`** (% clay, 0–5 cm mean from SoilGrids
250m).

```{r show-clay}
e <- R$effects$soilgrids_clay
cat("Adjustment set:", paste(e$adjustment, collapse = ", "), "\n")
cat(sprintf("  LM  naive = %+.4f g/kg per %%   -> direct = %+.4f  boot = [%+.4f, %+.4f]\n",
            e$lm$effect_naive, e$lm$effect,
            e$lm$effect_boot_ci[1L], e$lm$effect_boot_ci[2L]))
cat(sprintf("  BART direct (at +IQR/2) = %+.4f  CI95 = [%+.4f, %+.4f]\n",
            e$bart$effect, e$bart$effect_ci[1L], e$bart$effect_ci[2L]))
```

**Interpretation.** The naive slope +0.53 g/kg per 1 % clay
**overstates** the causal effect by a factor of **~2.7** once we
adjust for slope (erosional mass movement) and texture partners
(sand, bulk density). The direct effect is still positive but ~0.2
g/kg per 1 % clay — the mineral-protection signal is real but
smaller than the raw correlation suggests. The BART non-linear
estimate agrees with LM (0.20 vs 0.19), suggesting the clay → SOC
relationship is essentially linear in the Cerrado topsoil range.

## 4. Side-by-side summary

```{r summary-table, echo = FALSE}
rows <- lapply(R$effects, function(e) {
  data.frame(
    exposure      = e$exposure,
    adjustment    = paste(e$adjustment, collapse = ", "),
    naive         = e$lm$effect_naive,
    direct_lm     = e$lm$effect,
    direct_bart   = if (!is.null(e$bart)) e$bart$effect else NA_real_,
    boot_lo       = e$lm$effect_boot_ci[1L],
    boot_hi       = e$lm$effect_boot_ci[2L],
    stringsAsFactors = FALSE
  )
})
tbl <- do.call(rbind, rows)
knitr::kable(
  tbl,
  col.names = c("exposure", "adjustment set", "naive slope",
                "direct (LM)", "direct (BART)",
                "boot 2.5 %", "boot 97.5 %"),
  digits = 4L,
  caption = "Backdoor-adjusted direct effects on real Cerrado topsoil SOC."
)
```

## 5. What this vignette buys you over v1.3 Pillar 1

1. **Real data**, not `br_cerrado` synthetic. 1 095 WoSIS profiles,
   56 covariates, CC-BY-4.0 — the same bundle the v1.3.1 case-study
   benchmark uses.
2. **Three identified causal effects** with concrete soil-formation
   interpretation: the climate forcing of MAP → SOC (~0.007 g/kg
   per mm), the land-use pathway of tree cover → SOC (~2.05 g/kg
   per % tree cover, **more than 2× the naive** estimate), and the
   mineral-protection mechanism of clay → SOC (~0.19 g/kg per %
   clay, **~2.7× *less* than the naive** estimate).
3. **Block-bootstrap intervals** that respect the spatial clustering
   of WoSIS profiles — the asymptotic `confint(lm)` interval is
   reported for comparison but the bootstrap is the defensible
   number for a Brazilian pedometric paper.
4. **Non-linear cross-check** via BART: every estimate is
   reproduced within the LM CI, suggesting the linear SOC-covariate
   structure is a reasonable approximation for Cerrado topsoil.

The concrete output — the side-by-side "naive vs identified direct
effect" table — is exactly the evidence a soil-mapping paper needs
to argue that a given covariate has a *causal* (not merely
associational) role. The three numbers above are directly
citable.

## References
