---
title: "Pilar 2 — Physics-Informed ML for Soil Depth Profiles"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 2 — Physics-Informed ML for Soil Depth Profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4.5
)
```

## Abstract

Classical Digital Soil Mapping harmonises horizon observations via
equal-area quadratic splines [@Bishop1999; @Malone2009] — a purely
mathematical object disconnected from pedogenesis. The **Pillar 2** of
`edaphos` replaces that spline with two Physics-Informed Machine
Learning (PIML) [@Raissi2019; @Karniadakis2021; @Willcox2021pinns]
models of the soil depth profile: a parametric pedogenetic Ordinary
Differential Equation (ODE) integrated by `deSolve`
[@Soetaert2010desolve], and a Neural ODE [@Chen2018neuralode] trained
by back-propagation through a fixed-step Runge-Kutta integrator on
`torch` [@PichlerRtorch]. Both models are demonstrated on the
`aqp::sp4` reference pedons and a hierarchical covariate-conditioned
extension closes the loop with Pillar 5.

## 1. Pedogenetic ODE

Let \(y(z)\) denote a soil property (e.g. clay fraction, SOC) at depth
\(z\ge 0\). The one-dimensional kinetic model of translocation
[@Jenny1941; @Minasny2016] assumes a depth-dependent relaxation
towards a parent-material asymptote \(y_{\infty}\) with rate
\(\lambda(z)\):
\[
\frac{dy}{dz}\;=\;-\lambda(z)\,\bigl(y(z)-y_{\infty}\bigr),\qquad
\lambda(z)=\lambda_{0}\,e^{-\mu z}.
\tag{1}
\]
Parameters \((\lambda_{0},\mu,y_{\infty},y_{0})\) carry explicit
physical meaning: \(\lambda_{0}\) is the surface pedogenetic rate;
\(\mu\) controls how quickly that rate decays with depth; \(y_{0}\) is
the surface intercept; and \(y_{\infty}\) is the parent-material
extrapolation. The implementation
[`piml_profile_fit()`][piml_profile_fit] integrates (1) via
`deSolve::lsoda` at every loss evaluation and minimises
\[
\mathcal{J}(\theta)\;=\;
\sum_{j}\bigl(y(z_{j};\theta)-y_{j}\bigr)^{2}
\;+\;\eta\|\theta\|_{2}^{2},\tag{2}
\]
with Nelder--Mead and Tikhonov regularisation \(\eta\).[^psn]

[^psn]: Physics is imposed by the *forward model itself* — the
optimiser cannot produce a profile that violates the exponential
approach to \(y_{\infty}\), irrespective of data. This is the
strong-PIML regime sensu [@Karniadakis2021].

## 2. Dataset — `aqp::sp4`

We use the ten reference pedons of `aqp::sp4` with horizon
concentrations of clay, sand, Ca, Mg, K, CEC and CF.

```{r setup, message=FALSE}
library(edaphos)
.torch_ok <- requireNamespace("torch", quietly = TRUE) &&
             isTRUE(tryCatch(torch::torch_is_installed(),
                             error = function(e) FALSE))
if (!requireNamespace("aqp", quietly = TRUE)) {
  knitr::knit_exit("Package `aqp` not installed — skipping vignette.")
}
data(sp4, package = "aqp")
sp4$depth <- (sp4$top + sp4$bottom) / 2
head(sp4[, c("id", "name", "top", "bottom", "depth", "clay", "K")])
unique(sp4$id)
```

## 3. Mono-pedon fit: clay translocation in `napa`

The `napa` pedon is a canonical argillic-horizon case: clay increases
with depth through illuviation then converges to a plateau in the
parent material. The ODE (1) captures that dynamics without prescribing
the sign of \(y_{\infty}-y_{0}\).

```{r single-fit}
napa <- subset(sp4, id == "napa")
napa[, c("name", "depth", "clay")]
fit_napa <- piml_profile_fit(
  depths = napa$depth,
  values = napa$clay
)
fit_napa
```

```{r predict-grid}
grid_z <- seq(0, max(napa$depth) + 10, by = 1)
pred_napa <- data.frame(depth = grid_z,
                         clay  = predict(fit_napa, grid_z))
```

```{r plot-single, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
ggplot() +
  geom_line(data = pred_napa, aes(clay, depth),
            linewidth = 1, colour = "steelblue") +
  geom_point(data = napa, aes(clay, depth),
             size = 3, colour = "firebrick") +
  scale_y_reverse() +
  labs(x = "Clay (%)", y = "Depth (cm)",
       title = "Napa — pedogenetic ODE vs. observed horizons") +
  theme_minimal(base_size = 12)
```

## 4. Multi-pedon diagnostics

Fitting the full `sp4` dataset recovers an interpretable table where
every row has a direct pedological meaning.

```{r group-fit}
fits <- piml_profile_fit_group(
  sp4, id = "id", depth = "depth", value = "clay"
)
tab <- do.call(rbind, lapply(names(fits), function(k) {
  p <- fits[[k]]$params
  data.frame(
    pedon   = k,
    lambda0 = signif(p$lambda0, 3),
    mu      = signif(p$mu,      3),
    y_inf   = signif(p$y_inf,   3),
    y0      = signif(p$y0,      3),
    rmse    = signif(fits[[k]]$rmse, 3)
  )
}))
tab
```

## 5. Non-physical baselines

Two baselines bound the improvement claim: the mean predictor and the
linear-in-depth regression. A meaningful PIML gain should come from
the physical constraint, *not* from extra parameters.

```{r baseline}
baseline_rmse <- function(sub, target) {
  y <- sub[[target]]
  c(mean   = sqrt(mean((y - mean(y))^2)),
    linear = sqrt(mean(stats::residuals(
                     stats::lm(y ~ sub$depth))^2)))
}
bench <- do.call(rbind, lapply(split(sp4, sp4$id), function(sub) {
  bl <- baseline_rmse(sub, "clay")
  data.frame(pedon = sub$id[1],
             piml  = fits[[sub$id[1]]]$rmse,
             mean_baseline = bl["mean"],
             linear_baseline = bl["linear"])
}))
bench
```

## 6. Neural ODE variant

For profiles where the mono-asymptotic prior is violated — illuviated
bulges, E horizons beneath A, buried paleosols — `edaphos` offers a
Neural ODE alternative [@Chen2018neuralode]:
\[
\frac{dy}{dz}=f_{\theta}(z,y),\qquad
f_{\theta}=\text{MLP}(z,y;\,\theta).
\tag{3}
\]
The forward map is a fixed-step RK4 integrator written directly in
`torch`, so training back-propagates the squared-error loss through
the whole trajectory:
\[
\mathcal{J}_{\text{NODE}}(\theta)
   \;=\;\sum_{j}\Bigl(
      \underbrace{\int_{0}^{z_{j}}f_{\theta}(z,y(z))\,dz
                  +y_{0}}_{\text{RK4 rollout}}
      \;-\;y_{j}\Bigr)^{2}.
\]

```{r neural-ode, eval=.torch_ok}
nn_fit <- piml_neural_ode_fit(
  depths  = napa$depth, values = napa$clay,
  hidden  = c(16L, 16L), epochs = 500L,
  seed    = 1L, verbose = FALSE
)
nn_fit
pred_nn <- data.frame(depth = grid_z,
                       clay  = predict(nn_fit, grid_z))
```

```{r plot-neural, eval=.torch_ok && requireNamespace("ggplot2", quietly = TRUE)}
ggplot() +
  geom_line(data = pred_napa, aes(clay, depth,
            colour = "Parametric ODE"), linewidth = 1) +
  geom_line(data = pred_nn,   aes(clay, depth,
            colour = "Neural ODE"),     linewidth = 1) +
  geom_point(data = napa, aes(clay, depth),
             size = 3, colour = "firebrick") +
  scale_y_reverse() +
  scale_colour_manual(values = c("Parametric ODE" = "steelblue",
                                  "Neural ODE"    = "darkgreen")) +
  labs(x = "Clay (%)", y = "Depth (cm)",
       colour = NULL,
       title = "Napa — parametric vs. Neural ODE") +
  theme_minimal(base_size = 12)
```

## 7. Pillar 2 × Pillar 5 — the physics gate

Given a PIML fit (parametric or neural), the helper
[`al_physics_gate_piml()`][al_physics_gate_piml] constructs a
rejection function: candidates whose predicted target falls outside the
envelope \([\min(y_{0},y_{\infty}),\,\max(y_{0},y_{\infty})]\) inflated
by a `safety_factor` are eliminated from the greedy Active-Learning
selection (see the Pillar 5 vignette).

```{r gate, eval=.torch_ok}
gate <- al_physics_gate_piml(fit_napa, safety_factor = 1.2)
cand_dummy <- data.frame(dummy = 1:4)
gate(cand_dummy, predicted_mean = c(10, 25, 200, -5))
```

A hierarchical extension ([`piml_hierarchical_fit()`][piml_hierarchical_fit])
jointly fits a covariate-conditioned Neural ODE across pedons,
delivering a **per-location** envelope that
[`al_physics_gate_piml_hierarchical()`][al_physics_gate_piml_hierarchical]
feeds back into Pillar 5.

## 8. Bayesian posterior over the ODE parameters

[`piml_profile_fit()`][piml_profile_fit] returns a single point
estimate. For any honest downstream use — propagating parametric
uncertainty into a SOC-stock prediction, tuning a Pillar 5
Active-Learning acquisition under correct epistemic variance,
reporting publishable parameter intervals — we need the **posterior**
\(p(\lambda_0, \mu, y_\infty, y_0 \mid \text{depths}, \text{values})\),
not just the MAP.

The function
[`piml_profile_fit_bayesian()`][piml_profile_fit_bayesian] returns
that posterior through two nested approximations:

1. **Laplace approximation** (default, O(ms) per pedon). Gaussian
   posterior whose mean is the MAP and whose covariance is the inverse
   observed Fisher information at the MAP [@Bishop2006PRML, §4.4].
   Accurate when the posterior is well-identified; every call
   pre-samples 2 000 draws from \(N(\text{MAP}, \Sigma)\) for
   predictive convenience.
2. **Adaptive random-walk Metropolis** [@Haario2001] (~seconds per
   pedon). Proposal covariance starts at the Laplace covariance,
   scaled by the Roberts–Gelman–Gilks optimal factor
   \((2.38)^2 / d\), and is updated online by Haario recursion after a
   warm-up period — so non-Gaussian or multimodal posteriors are
   captured faithfully where Laplace would paper over them.

```{r bayes-laplace, eval=TRUE}
library(edaphos)

depths <- c(5, 15, 30, 60, 100)
values <- c(25, 18, 12, 8, 6.5)

fit_bayes <- piml_profile_fit_bayesian(depths, values,
                                         method = "laplace", seed = 1L)
fit_bayes
```

```{r bayes-predictive, eval=TRUE}
predict(fit_bayes,
         newdepths         = c(10, 20, 40, 80),
         interval          = 0.95,
         include_obs_noise = TRUE,
         seed              = 1L)
```

`include_obs_noise = TRUE` adds Gaussian observation noise with the
MAP-estimated \(\sigma\) onto every predictive draw, so the returned
interval represents the predictive distribution of a *future
observation* rather than the uncertainty on the mean function alone.

For the Neural-ODE variant the variational analogue is a **deep
ensemble** [@Lakshminarayanan2017ensemble]: `K` independent networks
trained from different random seeds produce an empirical posterior
whose mean and variance approximate the Bayesian predictive posterior
[@Wilson2020bayesian].

```{r bayes-deep-ensemble, eval=.torch_ok}
ens <- piml_neural_ode_fit_ensemble(depths, values,
                                       K = 5L, epochs = 300L, seed = 1L)
predict(ens, newdepths = c(10, 20, 40, 80), interval = 0.9)
```

## 9. Discussion and roadmap

The parametric ODE is mono-asymptotic and therefore optimal for
properties that decay monotonically toward a parent-material value
(SOC, some nutrients) but suboptimal for bimodal profiles, which the
Neural ODE handles by construction. Three complementary directions
remain on the roadmap:

* **Mass-conservation soft constraints** — augment the Neural ODE loss
  with a term penalising deviations from an integrated mass balance
  [@Minasny2017].
* **Adjoint-method Neural ODE** — for pedons deeper than 2 m, the
  back-propagation graph becomes memory-bound; the adjoint method of
  [@Chen2018neuralode] eliminates that scaling.
* **Hierarchical Bayesian pooling** — extending
  [`piml_hierarchical_fit()`][piml_hierarchical_fit] to the posterior
  regime that
  [`piml_profile_fit_bayesian()`][piml_profile_fit_bayesian] already
  exposes for single pedons, with covariate-conditioned priors over
  \((\lambda_0, \mu, y_\infty, y_0)\).

## References
