---
title: "Pilar 3 — 4D Pedometry: Stacked ConvLSTM Forecasts of SOC"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 3 — 4D Pedometry: Stacked ConvLSTM Forecasts of SOC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4.2
)
```

## Abstract

Most digital soil maps report a time-invariant property field, which
ignores the evidence that topsoil Soil Organic Carbon (SOC) responds
measurably to climate forcing on monthly to annual scales
[@Lehmann2015; @Minasny2017]. The **Pillar 3** of `edaphos` addresses
that gap with a stacked Convolutional LSTM [@Shi2015] trained in
sequence-to-sequence mode plus a multi-step rollout wrapper for
forward forecasting under known future drivers. A physics-informed
mass-balance regulariser optionally penalises violations of an
analytical SOC kinetic, fusing Pillar 2 into Pillar 3
[@Raissi2019; @ReichsteinDL2019].

## 1. From 3D to 4D pedometry

Let \(y(\mathbf{s},t)\) denote a topsoil property at location
\(\mathbf{s}\) and time \(t\in\{1,\ldots,T\}\). Traditional DSM
[@McBratney2003] estimates \(y(\mathbf{s})=\bar y(\mathbf{s},\cdot)\)
by collapsing over time. A 4D model retains the time dimension and
predicts the full spatio-temporal field
\(y(\mathbf{s},t)\mid \mathbf{X}(\mathbf{s},t^{\prime}\le t)\), where
\(\mathbf{X}\) is the driver stack (climate, vegetation, static
topography).

The **Convolutional LSTM cell** [@Shi2015] operationalises spatial
memory: at each time step the hidden state \(\mathbf{H}_{t}\) and the
cell state \(\mathbf{C}_{t}\) are tensors of the same spatial size as
the input, so memory propagates *with* its location:
\[
\begin{aligned}
\mathbf{i}_{t}&=\sigma(W_{xi}*\mathbf{X}_{t}+W_{hi}*\mathbf{H}_{t-1}+b_{i}),\\
\mathbf{f}_{t}&=\sigma(W_{xf}*\mathbf{X}_{t}+W_{hf}*\mathbf{H}_{t-1}+b_{f}),\\
\mathbf{g}_{t}&=\tanh(W_{xg}*\mathbf{X}_{t}+W_{hg}*\mathbf{H}_{t-1}+b_{g}),\\
\mathbf{o}_{t}&=\sigma(W_{xo}*\mathbf{X}_{t}+W_{ho}*\mathbf{H}_{t-1}+b_{o}),\\
\mathbf{C}_{t}&=\mathbf{f}_{t}\odot\mathbf{C}_{t-1}+\mathbf{i}_{t}\odot\mathbf{g}_{t},\\
\mathbf{H}_{t}&=\mathbf{o}_{t}\odot\tanh(\mathbf{C}_{t}),
\end{aligned}
\tag{1}
\]
with `*` a 2-D convolution and \(\odot\) the Hadamard product.
Stacking \(L\) cells feeds \(\mathbf{H}_{t}^{(\ell)}\) as the input to
layer \(\ell+1\), yielding a hierarchy of spatial receptive fields.

## 2. Synthetic SOC dynamics cube

To keep the vignette self-contained and reproducible,
[`temporal_synth_soc_cube()`][temporal_synth_soc_cube] integrates the
driver-response kinetic
\[
\mathrm{SOC}_{t+1}=\mathrm{SOC}_{t}+k_{\text{in}}P_{t}
                   -k_{\text{out}}\,\mathrm{SOC}_{t}\,P_{t}/\bar P
                   +\varepsilon,
\tag{2}
\]
with \(P_{t}\) the monthly precipitation field and \(\bar P\) its
long-term mean. The numerator \(k_{\text{in}}P_{t}\) models organic
input proportional to wet-season biomass turnover; the denominator
\(k_{\text{out}}\mathrm{SOC}\,P/\bar P\) captures humidity-modulated
decomposition [@Lehmann2015; @Minasny2017].

```{r setup, message=FALSE}
library(edaphos)
.torch_ok <- requireNamespace("torch", quietly = TRUE) &&
             isTRUE(tryCatch(torch::torch_is_installed(),
                             error = function(e) FALSE))
if (!.torch_ok) {
  knitr::knit_exit(
    "torch runtime (libtorch) not available — skipping vignette."
  )
}
cube <- temporal_synth_soc_cube(H = 12L, W = 12L, T_total = 18L,
                                seed = 7L)
str(cube)
```

```{r plot-cube, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
to_df <- function(mat, month) {
  dd <- expand.grid(row = seq_len(nrow(mat)), col = seq_len(ncol(mat)))
  dd$soc <- as.vector(mat); dd$month <- month
  dd
}
soc_df <- rbind(to_df(cube$soc[1L, , ],  "month 1"),
                 to_df(cube$soc[9L, , ],  "month 9"),
                 to_df(cube$soc[18L, , ], "month 18"))
ggplot(soc_df, aes(col, -row, fill = soc)) +
  geom_raster() + facet_wrap(~ month) +
  scale_fill_viridis_c(name = expression("SOC ("*g~kg^{-1}*")")) +
  coord_equal() + theme_minimal(base_size = 12) +
  labs(x = NULL, y = NULL,
       title = "Spatio-temporal SOC cube (18 months)")
```

## 3. Train / forecast split

We reserve the last six months for an out-of-sample forecast evaluation
and train on the first twelve.

```{r split}
past_tensor   <- temporal_cube_to_tensor(cube, t_slice = 1:12)
future_tensor <- temporal_cube_to_tensor(cube, t_slice = 13:18)
dim(past_tensor$sequence)
dim(past_tensor$target)
```

## 4. Stacked ConvLSTM in sequence-to-sequence mode

```{r train, message=FALSE}
fit <- temporal_convlstm_fit(
  sequence        = past_tensor$sequence,
  target          = past_tensor$target,
  hidden_dims     = c(12L, 6L),
  kernel_size     = 3L,
  return_sequence = TRUE,
  epochs          = 120L, lr = 0.02,
  seed            = 1L, verbose = FALSE
)
fit
```

```{r loss-curve, eval=requireNamespace("ggplot2", quietly = TRUE)}
loss_df <- data.frame(epoch = seq_along(fit$loss_history),
                       loss  = fit$loss_history)
ggplot(loss_df, aes(epoch, loss)) +
  geom_line(linewidth = 0.7) + scale_y_log10() +
  theme_minimal(base_size = 12) +
  labs(title = "Training loss — stacked ConvLSTM (2 layers)",
       x = "Epoch", y = "MSE (log scale)")
```

## 5. Multi-step rollout forecast

When future driver channels are known (e.g. climate forecasts
[@ReichsteinDL2019]), a single forward pass of the trained ConvLSTM on
the concatenated past + future window returns the per-step prediction.

```{r rollout}
forecast <- temporal_convlstm_rollout(
  fit,
  past_sequence  = past_tensor$sequence,
  future_drivers = future_tensor$sequence
)
dim(forecast)
```

```{r rmse-map}
tgt_norm  <- future_tensor$target[1L, , , ]
pred_norm <- forecast[1L, , , ]
rmse_month <- sapply(seq_len(6), function(t)
  sqrt(mean((pred_norm[t, , ] - tgt_norm[t, , ])^2)))
data.frame(forecast_month = 13:18, rmse_norm = signif(rmse_month, 3))
per_pixel <- sqrt(apply((pred_norm - tgt_norm)^2, c(2, 3), mean))
summary(as.vector(per_pixel))
```

```{r plot-forecast-map, eval=requireNamespace("ggplot2", quietly = TRUE)}
df <- rbind(
  to_df(future_tensor$scaling$soc_mu +
          future_tensor$scaling$soc_sd * tgt_norm[1L, , ],
        "month 13 — observed"),
  to_df(future_tensor$scaling$soc_mu +
          future_tensor$scaling$soc_sd * pred_norm[1L, , ],
        "month 13 — forecast"),
  to_df(future_tensor$scaling$soc_mu +
          future_tensor$scaling$soc_sd * tgt_norm[6L, , ],
        "month 18 — observed"),
  to_df(future_tensor$scaling$soc_mu +
          future_tensor$scaling$soc_sd * pred_norm[6L, , ],
        "month 18 — forecast")
)
df$month <- factor(df$month, levels = c(
  "month 13 — observed", "month 13 — forecast",
  "month 18 — observed", "month 18 — forecast"))
ggplot(df, aes(col, -row, fill = soc)) +
  geom_raster() + facet_wrap(~ month, nrow = 2) +
  scale_fill_viridis_c(name = expression("SOC ("*g~kg^{-1}*")")) +
  coord_equal() + theme_minimal(base_size = 12) +
  labs(x = NULL, y = NULL,
       title = "Forecast vs. ground truth, months 13 and 18")
```

## 6. Physics-informed ConvLSTM

For Pillars 2 × 3 fusion, a mass-balance prior of the form (2) can be
injected into the loss as a soft constraint
[@Raissi2019; @Karniadakis2021; @Willcox2021pinns]:
\[
\mathcal{L}=\underbrace{\mathrm{MSE}(\hat y,y)}_{\text{data}}
           +\lambda_{\text{phys}}\,
           \mathrm{MSE}\!\bigl(\Delta\hat y_{t},
               k_{\text{in}}P_{t}-k_{\text{out}}\hat y_{t}P_{t}/\bar P\bigr).
\]
In [`temporal_convlstm_fit()`][temporal_convlstm_fit] this is enabled
by setting `physics_lambda > 0`. Two auxiliary traces —
`loss_fit_history` and `loss_phys_history` — are stored so the data and
physics contributions can be monitored independently during training.

## 7. Discussion

Pillar 3 establishes the scaffold for a genuinely 4-D DSM pipeline —
distinct from repeated 3-D snapshots — and connects it back to Pillar
2 through the shared kinetic (2). Production-scale use cases require
real spatio-temporal stacks (MODIS, ERA5, MapBiomas), multi-step
autoregressive rollout, and hysteresis regularisers; those are
scheduled for future releases.

## References
