---
title: "Pilar 4 — Contrastive Raster Embeddings as Active-Learning Covariates"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 4 — Contrastive Raster Embeddings as Active-Learning Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4.2
)
```

## Abstract

Self-Supervised Learning (SSL) has emerged as the dominant
representation-learning paradigm in computer vision
[@Chen2020simclr; @He2020moco; @Oord2018cpc] and is increasingly
applied to remote sensing [@Jean2019tile2vec; @ReichsteinDL2019]. The
**Pillar 4** of `edaphos` ships a minimal SimCLR-style pipeline
[@Chen2020simclr] that pre-trains a small raster encoder on unlabelled
covariate patches; the resulting per-pixel embeddings are then used as
additional covariates inside the Pillar 5 Active Learning loop on
`br_cerrado`. We report a head-to-head comparison of the identical AL
policy with and without the learned features.

## 1. Contrastive representation learning

Given a mini-batch \(\{\mathbf{x}_{i}\}_{i=1}^{B}\) of raster patches,
SimCLR draws two independent stochastic augmentations
\(\tilde{\mathbf{x}}_{i}^{(1)},\tilde{\mathbf{x}}_{i}^{(2)}\) of each
patch, encodes them with a shared CNN \(f_{\theta}\) and a projection
head \(g_{\phi}\), and minimises the normalised temperature-scaled
cross-entropy (NT-Xent) loss [@Chen2020simclr]:
\[
\mathcal{L}_{\text{NT-Xent}}
   \;=\;-\frac{1}{2B}
         \sum_{i=1}^{B}\sum_{v\in\{1,2\}}
           \log\frac{\exp(\mathrm{sim}(\mathbf{z}_{i}^{(v)},
                                         \mathbf{z}_{i}^{(v^{\prime})})/\tau)}
           {\sum_{k\ne(i,v)}\exp(\mathrm{sim}(\mathbf{z}_{i}^{(v)},
                                              \mathbf{z}_{k})/\tau)},
\]
where \(\mathbf{z}=g_{\phi}(f_{\theta}(\tilde{\mathbf{x}}))\),
\(\mathrm{sim}(\cdot,\cdot)\) is cosine similarity and \(\tau\) is the
temperature. After pre-training, the projection head is discarded and
the *backbone* feature vector \(f_{\theta}(\mathbf{x})\) is used as the
reusable representation.

## 2. Data preparation

Each of the 2025 pixels of `br_cerrado` is made the centre of a
\(7\times 7\times 5\) patch with five normalised channels (elevation,
slope, TWI, mean annual precipitation, NDVI). Reflection padding
preserves the original grid size.

```{r setup, message=FALSE}
library(edaphos)
.torch_ok <- requireNamespace("torch", quietly = TRUE) &&
             isTRUE(tryCatch(torch::torch_is_installed(),
                             error = function(e) FALSE))
if (!.torch_ok) {
  knitr::knit_exit(
    "torch runtime (libtorch) not available — skipping vignette."
  )
}
data(br_cerrado, package = "edaphos")

covs_base <- c("elev", "slope", "twi", "map_mm", "ndvi")
H <- length(unique(br_cerrado$y))
W <- length(unique(br_cerrado$x))
stopifnot(H * W == nrow(br_cerrado))
c(H = H, W = W)
```

```{r patchify}
grid <- br_cerrado[order(br_cerrado$y, br_cerrado$x), ]
mat <- function(col) matrix(grid[[col]], nrow = H, ncol = W, byrow = TRUE)
z_mat <- function(m) (m - mean(m)) / max(stats::sd(m), 1e-6)
channels <- lapply(covs_base, function(cc) z_mat(mat(cc)))
cube <- array(unlist(channels), dim = c(H, W, length(covs_base)))
cube <- aperm(cube, c(3L, 1L, 2L))
mirror_pad <- function(m, p) {
  top <- m[p:1, , drop = FALSE]
  bot <- m[nrow(m):(nrow(m) - p + 1L), , drop = FALSE]
  tall <- rbind(top, m, bot)
  left  <- tall[, p:1, drop = FALSE]
  right <- tall[, ncol(tall):(ncol(tall) - p + 1L), drop = FALSE]
  cbind(left, tall, right)
}
patch_size <- 7L
pad <- (patch_size - 1L) %/% 2L
C <- dim(cube)[1L]
padded <- array(0, dim = c(C, H + 2L * pad, W + 2L * pad))
for (cc in seq_len(C)) padded[cc, , ] <- mirror_pad(cube[cc, , ], pad)
patches <- array(0, dim = c(H * W, C, patch_size, patch_size))
k <- 1L
for (i in seq_len(H)) for (j in seq_len(W)) {
  patches[k, , , ] <- padded[, i:(i + 2L * pad), j:(j + 2L * pad)]
  k <- k + 1L
}
dim(patches)
```

## 3. SimCLR pre-training

Four hundred randomly-drawn patches are pre-trained for 60 epochs on
Adam. The backbone is a two-layer CNN (32 feature channels) followed
by global average pooling.

```{r pretrain, message=FALSE}
set.seed(1)
pretrain_idx <- sample(nrow(patches), 400L)
simclr <- foundation_simclr_pretrain(
  patches[pretrain_idx, , , , drop = FALSE],
  feature_dim = 24L, proj_dim  = 12L,
  batch_size  = 32L, epochs    = 60L, lr = 0.005,
  temperature = 0.2, noise_sd  = 0.1,
  seed        = 1L, verbose    = FALSE
)
simclr
```

## 4. Per-pixel embedding extraction

```{r embed}
emb <- foundation_simclr_embed(simclr, patches)
dim(emb)
colnames(emb) <- paste0("emb", sprintf("%02d", seq_len(ncol(emb))))
br_ext <- cbind(grid, emb)
head(br_ext[, c(covs_base, "soc", head(colnames(emb), 3L))])
```

## 5. Controlled experiment — AL with and without SimCLR features

Same cLHS seed, same budget, same hybrid policy; only the covariate
list changes. Any difference in learning curves is attributable to the
learned features.

```{r al-compare, message=FALSE}
set.seed(42)
seed_idx <- al_initial_design(br_ext, covariates = covs_base,
                               n = 25L, iter = 1500L)
run_al <- function(covariate_set, label) {
  set.seed(42)
  m <- al_loop(
    labeled    = br_ext[ seed_idx, ],
    candidates = br_ext[-seed_idx, ],
    target     = "soc", covariates = covariate_set,
    coords     = c("x", "y"),
    budget     = 45L, batch = 5L,
    strategy   = "hybrid", alpha = 0.7,
    num.trees  = 500L, verbose = FALSE
  )
  cbind(al_history(m), setup = label)
}
covs_ext <- c(covs_base, colnames(emb))
h_base <- run_al(covs_base, "baseline (5 covariates)")
h_ext  <- run_al(covs_ext,  "baseline + SimCLR (29 covariates)")
h <- rbind(h_base, h_ext)
tail(h[h$setup == "baseline (5 covariates)", ], 4)
tail(h[h$setup == "baseline + SimCLR (29 covariates)", ], 4)
```

```{r plot-compare, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
ggplot(h, aes(n_labeled, rmse_oob, colour = setup)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  labs(x = "Labelled sample size n_t",
       y = expression("OOB RMSE ("*g~kg^{-1}*" SOC)"),
       title = "AL on br_cerrado — SimCLR features as auxiliary covariates",
       colour = "Feature set") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 6. Discussion

On this small synthetic grid (N = 2025 pixels, five physically-coded
covariates, 60 SimCLR epochs) the learned features typically yield a
marginal OOB-RMSE improvement, which is itself informative: the
encoder extracts signal from the covariate stack **with zero labels
consumed during pre-training** [@Jean2019tile2vec].

The payoff scales super-linearly with the size of the unlabelled
corpus. Under a realistic Cerrado-wide pre-training with optical + SAR
+ DEM + climate inputs, the learned representation captures transfer
structure unavailable to any regional model [@Chen2020simclr; @ReichsteinDL2019],
and becomes the de-facto feature layer for Pillars 2, 3 and 5. That
planetary-scale workflow is scheduled for the next major release.

## 7. MoCo v2 — upgraded momentum-contrast encoder

The SimCLR scaffold compares every anchor against `2B - 2` negatives
drawn from the current mini-batch. That design is simple but couples
the number of negatives to the mini-batch size, which is the primary
reason production SSL pipelines since 2020 have switched to
**Momentum Contrast (MoCo)** [@He2020moco; @Chen2020moco2].
The edaphos MoCo v2 implementation
[`foundation_moco_pretrain()`][foundation_moco_pretrain] ships three
upgrades over the SimCLR scaffold:

1. **Momentum key encoder** updated by exponential moving average
   \(\theta_k \leftarrow m\,\theta_k + (1-m)\,\theta_q\) — keys are
   stable across batches.
2. **Dictionary queue** of past keys — every anchor sees
   \(K \gg B\) negatives with no additional batch-size cost.
3. **Wider residual backbone + 2-layer projection head with
   BatchNorm** — closer to the MoCo v2 recipe than the SimCLR
   scaffold's shallow CNN.

Augmentations are tailored to multi-channel raster patches: spatial
random resized crop, horizontal / vertical flip, 90-degree rotations,
per-channel Bernoulli dropout (missing-band simulation), spatial
cutout (cloud-mask simulation), per-channel multiplicative
brightness jitter, and additive Gaussian sensor noise.

```{r moco-pretrain, eval=.torch_ok, message=FALSE}
set.seed(1)
moco_fit <- foundation_moco_pretrain(
  patches[pretrain_idx, , , , drop = FALSE],
  feature_dim = 32L, proj_dim    = 16L,
  queue_size  = 128L, batch_size = 32L,
  momentum    = 0.99,
  temperature = 0.1,
  epochs = 30L, lr = 0.01,
  seed   = 1L, verbose = FALSE
)
moco_fit
```

```{r moco-loss, eval=.torch_ok && requireNamespace("ggplot2", quietly = TRUE)}
loss_df <- rbind(
  data.frame(epoch = seq_along(simclr$loss_history),
             loss  = simclr$loss_history, method = "SimCLR"),
  data.frame(epoch = seq_along(moco_fit$loss_history),
             loss  = moco_fit$loss_history, method = "MoCo v2")
)
ggplot(loss_df, aes(epoch, loss, colour = method)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.4) +
  labs(x = "Epoch", y = "contrastive loss",
       title = "SimCLR vs. MoCo v2 on the same br_cerrado patches",
       colour = "Method") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The MoCo v2 loss uses the same InfoNCE form but against a much larger
negative pool, so its absolute scale differs from SimCLR's NT-Xent and
the curves should be compared by their relative downward slope rather
than their absolute values.

## 8. MoCo embeddings as Pillar 5 covariates

The final step is identical to the SimCLR workflow — extract backbone
features and pipe them into the Active-Learning loop:

```{r moco-embed, eval=.torch_ok}
moco_emb <- foundation_moco_embed(moco_fit, patches)
colnames(moco_emb) <- paste0("mocoemb", sprintf("%02d", seq_len(ncol(moco_emb))))
br_moco <- cbind(grid, moco_emb)
```

The `moco_emb` columns can be inspected for spatial coherence (nearby
pixels should have similar embeddings if the encoder has learned
anything meaningful about the covariate topology) and added to
`covariates =` in `al_loop()` alongside or instead of the SimCLR
columns.

## 9. Planetary-scale pre-training on real multi-source mosaics

The in-memory `(N, C, H, W)` array used in sections 3-8 works for
classroom-sized demos but does not scale: a continental SoilGrids +
WorldClim + SRTM + MODIS + ERA5 mosaic at 250 m resolution easily
exceeds any laptop's RAM. Pillar 4 therefore ships a **lazy
tile-streaming pipeline** that reads patches on the fly from a
`terra::SpatRaster` mosaic.

### 9.1 Tile sources

Five public entry points cover the canonical DSM stack:

```{r tile-sources-doc, eval = FALSE}
# Keyless public sources -- no account required.
foundation_tile_source_soilgrids(variables = c("soc","clay","phh2o"),
                                   depth = "0-5cm",
                                   aoi   = c(-50, -46, -16, -12))
foundation_tile_source_worldclim(variables = c("prec","tavg"),
                                   country   = "BRA",
                                   aoi = c(-50, -46, -16, -12))
foundation_tile_source_srtm(aoi = c(-50, -46, -16, -12),
                              derive = c("slope", "aspect"))

# Credential-gated sources (documented stubs).
foundation_tile_source_modis(...)   # NASA EarthData login
foundation_tile_source_era5(...)    # Copernicus CDS API key
```

The MODIS and ERA5 stubs fail with a precise install / authentication
hint so users can plug their own pre-processed mosaics straight into
[`foundation_tile_align()`][foundation_tile_align] without modifying
the downstream code.

### 9.2 Alignment to a common grid

```{r tile-align-doc, eval = FALSE}
stack <- foundation_tile_align(
  sources = list(
    soil    = foundation_tile_source_soilgrids(
                 c("soc","clay"), aoi = aoi),
    climate = foundation_tile_source_worldclim(
                 c("prec","tavg"), country = "BRA", aoi = aoi),
    relief  = foundation_tile_source_srtm(aoi, derive = "slope")
  ),
  target_res = 0.0025,          # ~250 m
  target_crs = "EPSG:4326",
  aoi        = aoi
)
stack          # SpatRaster (n_rows, n_cols, n_layers)
```

### 9.3 Lazy patch dataset

```{r tile-dataset-doc, eval=requireNamespace("terra", quietly = TRUE)}
# For vignette reproducibility, build a synthetic 4-channel 48x48
# raster that stands in for the real SoilGrids + WorldClim + SRTM
# stack above. Replace with the `stack` from 9.2 on a real AoI.
library(edaphos)
library(terra)
set.seed(1)
r <- rast(ncols = 48, nrows = 48,
          xmin = -50, xmax = -46, ymin = -16, ymax = -12,
          crs = "EPSG:4326")
layers <- lapply(seq_len(4), function(k) {
  rr <- r
  values(rr) <- 10 * k + stats::rnorm(ncell(r), sd = k)
  names(rr) <- paste0("ch", k)
  rr
})
stack <- Reduce(function(a, b) c(a, b), layers)

dataset <- foundation_tile_dataset(
  stack,
  patch_size = 8L,
  n_patches  = 200L,
  seed       = 1L
)
dataset
```

### 9.4 MoCo v2 training on the streamed dataset

```{r tile-train, eval=requireNamespace("torch", quietly = TRUE) && reticulate::py_available() || .torch_ok}
ck_dir <- tempfile("moco_ck_")
moco_tiles <- foundation_moco_pretrain_tiles(
  dataset,
  feature_dim = 32L, proj_dim = 16L,
  queue_size  = 64L, batch_size = 16L,
  epochs = 20L, lr = 0.02,
  seed   = 1L, verbose = FALSE,
  checkpoint_dir   = ck_dir,
  checkpoint_every = 10L
)
moco_tiles
```

Checkpointing writes `encoder_q.pt`, `encoder_k.pt`, `queue.pt` and
`state.rds` to `checkpoint_dir`. Restart with `resume = ck_dir` after
any number of restarts without losing progress.

### 9.5 Embedding the full mosaic

```{r tile-embed, eval=.torch_ok}
emb <- foundation_moco_embed_raster(
  moco_tiles,
  stack   = stack,
  dataset = dataset,
  patch_size = 8L,
  stride     = 8L
)
emb
```

The output is a `terra::SpatRaster` with one band per backbone
feature (`feature_dim = 32` here). For visualisation we project the
embedding onto its top-3 principal components and render them as an
RGB composite:

```{r tile-embed-plot, eval=.torch_ok && requireNamespace("ggplot2", quietly = TRUE)}
emb_mat <- terra::values(emb)
keep    <- stats::complete.cases(emb_mat)
pr      <- stats::prcomp(emb_mat[keep, , drop = FALSE], scale. = TRUE)
rgb_vals <- matrix(NA_real_, nrow = nrow(emb_mat), ncol = 3L)
rgb_vals[keep, ] <- pr$x[, 1:3]
rgb_vals <- apply(rgb_vals, 2L, function(v) {
  v <- (v - min(v, na.rm = TRUE)) /
       diff(range(v, na.rm = TRUE))
  v
})
rgb_rast <- terra::rast(emb, nlyrs = 3L)
terra::values(rgb_rast) <- rgb_vals
names(rgb_rast) <- c("PC1", "PC2", "PC3")
terra::plotRGB(rgb_rast, r = 1L, g = 2L, b = 3L, stretch = "lin",
               axes = TRUE, main = "MoCo v2 embedding projected to PC1-3")
```

### 9.6 Scaling to continental mosaics

The same recipe scales to tens of millions of patches on production
hardware: swap `dataset` for one built on a continental VRT
(`terra::vrt()` stitches arbitrarily many tiles), raise
`batch_size`, `queue_size`, and `epochs`, and run the loop on a GPU
by setting `device = "mps"` (Apple Silicon) or `device = "cuda"`
(NVIDIA), both supported by
[`foundation_moco_pretrain_tiles()`][foundation_moco_pretrain_tiles]
as of v1.2.0. The checkpoint protocol is the same, so overnight or
multi-day runs can be resumed without state loss.

## 10. Downstream fine-tuning and pretrained weight distribution

Everything above produces a self-supervised *representation*. Turning
that representation into a supervised prediction — classifying a
patch by soil order, regressing a patch into an SOC point estimate —
needs a task-specific head on top of the encoder and a labelled
dataset to train it.

v1.2.0 adds two entry points,
[`foundation_fit_classifier()`][foundation_fit_classifier] and
[`foundation_fit_regressor()`][foundation_fit_regressor], that wrap
head construction, training loop, validation split and target
normalisation behind a single function call. Both support two
regimes from the transfer-learning literature [@He2019rethinking;
@Kornblith2019transfer]:

* **Linear probe** (`freeze_backbone = TRUE`, `head = "linear"`) —
  the encoder is frozen; only a single `nn_linear(feature_dim,
  n_out)` head is trained. Canonical benchmark for self-supervised
  representations as fixed feature extractors.
* **Full fine-tuning** (`freeze_backbone = FALSE`) — encoder and
  head trained jointly with a two-group learning rate
  (`lr * backbone_lr_mult` for the backbone, `lr` for the head).
  Usually the better option once the downstream dataset exceeds
  ~500 patches per class.

```{r finetune-classifier, eval=FALSE}
soil_order <- factor(sample(c("Oxisol", "Ultisol", "Inceptisol"),
                              200L, replace = TRUE))
fit <- foundation_fit_classifier(
  encoder         = moco,
  x               = labelled_patches,
  y               = soil_order,
  freeze_backbone = TRUE, head = "linear",
  epochs = 40L, lr = 1e-3, val_split = 0.2,
  device = "mps", seed = 1L, verbose = TRUE
)
predict(fit, labelled_patches[1:5, , , , drop = FALSE], type = "prob")
```

For a continuous target the regressor handles target normalisation
internally and offers a Huber loss that is robust to pedogenic
outliers:

```{r finetune-regressor, eval=FALSE}
fit_soc <- foundation_fit_regressor(
  encoder = moco, x = labelled_patches, y = soc_g_per_kg,
  freeze_backbone = FALSE, head = "mlp", hidden = c(64L, 32L),
  epochs = 80L, lr = 1e-3, loss = "huber",
  val_split = 0.2, device = "mps", seed = 1L
)
predict(fit_soc, new_patches)
```

### 10.1 Published Cerrado MoCo v2 encoder on Zenodo

Self-supervised pretraining on a continental stack is heavy — hours
of GPU time, gigabytes of raster downloads. Rather than asking every
user to reproduce that from scratch, `edaphos` **hosts its
pretrained encoders on Zenodo** under a CC-BY-4.0 licence and ships
a registry-based download layer:

- [`foundation_weights_list()`][foundation_weights_list] — catalogue
  of every published encoder.
- [`foundation_weights_download()`][foundation_weights_download] —
  fetches from Zenodo, verifies SHA-256, caches under
  `tools::R_user_dir("edaphos")`.
- [`foundation_weights_load()`][foundation_weights_load] — rebuilds
  the in-memory `edaphos_foundation_moco` wrapper with the state
  dict loaded and shape metadata populated.

The first published encoder, **`edaphos-cerrado-moco-v1`**, was
pretrained on an Apple M1 Max over 50 000 16×16 tiles of a core
Cerrado AoI (longitude −53 to −43, latitude −23 to −10). Channels
are a multi-source stack aligned to a 0.01 degree (~1 km) grid:

| Source                              | Channels                                 |
|:---|:---|
| SoilGrids 250m (0–5 cm mean)        | SOC, clay, sand, pH(H₂O), bulk density   |
| WorldClim 2.1 (Brazil country pack) | 12 monthly precipitation + 12 monthly mean temperature |
| SRTM 30 arc-second                  | elevation, slope                         |

```{r pretrained-load, eval=FALSE}
moco_cerrado <- foundation_weights_load("edaphos-cerrado-moco-v1",
                                          verbose = TRUE)
emb <- foundation_moco_embed(moco_cerrado, my_patches)
fit <- foundation_fit_classifier(moco_cerrado, my_patches,
                                   soil_order, epochs = 30L,
                                   device = "mps")
```

Custom encoders produced by the user's own
[`foundation_moco_pretrain_tiles()`][foundation_moco_pretrain_tiles]
run consume the same API — pass the local `.pt` path instead of the
registry name:

```{r pretrained-custom, eval=FALSE}
my_moco <- foundation_weights_load(
  "path/to/my_encoder.pt",
  n_channels = 9L, feature_dim = 64L, proj_dim = 32L
)
```

## References
