---
title: "Pilar 5 — Active Learning on a Brazilian Cerrado Transect"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 5 — Active Learning on a Brazilian Cerrado Transect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4.5
)
```

## Abstract

We apply the Pillar 5 Active Learning loop of `edaphos` to the
bundled `br_cerrado` sample — a reproducible 45 × 45 pixel synthetic
surrogate for a 30 km square near Brasília (Distrito Federal, Brazil).
The exercise mirrors the realistic cost structure of a DSM project in
the Cerrado biome [@Arrouays2014globalsoilmap; @Minasny2017]: limited
sampling budget, non-trivial covariate interactions, and a soil
organic carbon response driven by topography, climate and vegetation
cover.

## 1. Bundled surrogate dataset

`br_cerrado` is generated by the script
`data-raw/prepare_br_cerrado.R`, which also includes a ready-to-run
live-fetch block against SoilGrids 250 m [@HenglWoSIS], WorldClim and
SRTM via the `geodata` package. Synthetic data are used here only to
ensure that the vignette builds offline, reproducibly, in every CI
environment.

```{r setup, message=FALSE}
library(edaphos)
data(br_cerrado, package = "edaphos")
str(br_cerrado)
summary(br_cerrado[, c("soc", "elev", "twi", "ndvi", "map_mm")])
```

```{r preview-map, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
ggplot(br_cerrado, aes(x, y, fill = soc)) +
  geom_raster() +
  scale_fill_viridis_c(option = "D",
                        name = expression("SOC ("*g~kg^{-1}*")")) +
  coord_equal() +
  labs(title = "Synthetic ground truth — topsoil SOC (0-5 cm)",
       x = "Longitude (°)", y = "Latitude (°)") +
  theme_minimal(base_size = 12)
```

## 2. Sampling scenario

Budget \(N=80\): \(n_{0}=25\) initial samples selected by cLHS
[@Minasny2006clhs] and \(N-n_{0}=55\) additional samples chosen by the
Pillar 5 hybrid policy (Section 1 of `pilar5-active-learning`).

```{r seed}
covs <- c("elev", "slope", "twi", "map_mm", "ndvi")
set.seed(101)
seed_idx   <- al_initial_design(br_cerrado, covariates = covs,
                                n = 25, iter = 1500)
labeled0   <- br_cerrado[ seed_idx, ]
candidates <- br_cerrado[-seed_idx, ]
c(n_labeled = nrow(labeled0), n_candidates = nrow(candidates))
```

## 3. Closed-loop run

```{r loop, message=FALSE}
set.seed(101)
model <- al_loop(
  labeled    = labeled0, candidates = candidates,
  target     = "soc", covariates = covs,
  coords     = c("x", "y"),
  budget     = 55, batch = 5,
  strategy   = "hybrid", alpha = 0.7,
  num.trees  = 500L, verbose = FALSE
)
model
```

## 4. Learning curve

```{r history}
h <- al_history(model)
h
```

```{r plot-learning, eval=requireNamespace("ggplot2", quietly = TRUE)}
ggplot(h, aes(n_labeled, rmse_oob)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  labs(x = "Labelled sample size n_t",
       y = expression("OOB RMSE ("*g~kg^{-1}*" SOC)"),
       title = "Cerrado recorte — hybrid policy learning curve") +
  theme_minimal(base_size = 12)
```

## 5. Where did the algorithm decide to sample?

Superimposing seed locations on the queried set visualises the policy's
preference for covariate-transition zones — the hallmark of uncertainty
sampling [@Settles2009; @Lagacherie2022].

```{r plot-sampling, eval=requireNamespace("ggplot2", quietly = TRUE)}
seed_xy <- labeled0[, c("x", "y")]; seed_xy$type <- "Seed (cLHS)"
queried_xy <- model$labeled[-seq_len(nrow(labeled0)), c("x", "y")]
queried_xy$type <- "AL-queried"
samp <- rbind(seed_xy, queried_xy)

ggplot() +
  geom_raster(data = br_cerrado, aes(x, y, fill = soc)) +
  scale_fill_viridis_c(option = "D",
                        name = expression("SOC ("*g~kg^{-1}*")")) +
  geom_point(data = samp, aes(x, y, shape = type, colour = type),
             size = 2.2, stroke = 1.1) +
  scale_colour_manual(values = c("Seed (cLHS)" = "white",
                                  "AL-queried" = "firebrick")) +
  scale_shape_manual(values  = c("Seed (cLHS)" = 1, "AL-queried" = 4)) +
  coord_equal() +
  labs(title = "Adaptive sampling on the Cerrado grid",
       x = "Longitude (°)", y = "Latitude (°)") +
  theme_minimal(base_size = 12)
```

## 6. Transitioning to live SoilGrids data

To re-run this vignette on live SoilGrids 250 m data
[@HenglWoSIS] plus WorldClim precipitation and SRTM elevation:

1. Open `data-raw/prepare_br_cerrado.R` and toggle `if (FALSE)` to
   `if (TRUE)` — the block fetches the real covariate stack for the
   same bounding box.
2. Optionally enrich with MapBiomas land-use/land-cover as a
   categorical covariate.
3. Re-save with `usethis::use_data(br_cerrado, overwrite = TRUE)`.
4. Re-knit this vignette unchanged.

## 7. Discussion

Two observations follow. First, cLHS alone already achieves a
respectable seed RMSE because it is designed to span the joint
covariate distribution [@Minasny2006clhs]. Second, the marginal gain
from Active Learning is largest in the transition zones where soil
processes change regime — precisely where expert judgement historically
placed transects. The formalism in Pillar 5 replaces that judgement
with a reproducible, model-driven criterion.

## References
