---
title: "Pilar 3 on real data — 4D NDVI forecasting with sequential Bayesian assimilation"
author: "Hugo Rodrigues"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pilar 3 on real data — 4D NDVI forecasting with sequential Bayesian assimilation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7.5, fig.height = 4.5,
  fig.align = "center", out.width = "92%", dpi = 150
)
```

## 1. Why this vignette exists

The original Pillar 3 vignette (`vignette("pilar3-4d-soc")`) demonstrates
the stacked ConvLSTM [@Shi2015] on the synthetic SOC cube produced by
`temporal_synth_soc_cube()`. As of **v1.5.0** the Pillar 3 story runs on
**real Cerrado spatio-temporal data**: 168 monthly time steps
(January 2010 – December 2023) of three channels assembled over a
2° × 2° AoI centred on the Goiás / Minas Gerais triple-junction
(lon –48 to –46, lat –16 to –14, 10 × 10 = 100 cells at 0.2° resolution):

1. **NDVI** from MODIS MOD13Q1 250 m 16-day composites aggregated to
   monthly means [@Didan2015mod13q1] — the forecast target.
2. **Monthly precipitation** (mm/month) from NASA POWER
   (MERRA-2 bias-corrected) monthly mean daily totals
   [@Stackhouse2018power; @Sparks2018nasapower]. We originally
   targeted CHIRPS 0.05° [@Funk2015chirps] but the Climate Hazards
   Group data portal was returning HTTP 403 for every v2.0
   `global_monthly/tifs/` request at the v1.5.0 freeze; POWER is
   a fully-open NASA alternative with comparable accuracy over the
   Cerrado and a single-point REST API that returns the full
   14-year series in one call.
3. **Mean monthly air temperature** (°C) from NASA POWER T2M
   [@Stackhouse2018power], fetched in the same REST call as the
   precipitation. (We originally targeted WorldClim 2.1 static
   climatology [@Fick2017worldclim] but the `geodata.ucdavis.edu`
   server was in maintenance at the v1.5.0 freeze; POWER T2M has
   the added benefit of being **year-specific** rather than a long-
   term average, so all three channels now carry genuine year-to-
   year climate variability.)

The vignette also demonstrates the v1.5.0 new function
[`temporal_kalman_update()`](../reference/temporal_kalman_update.html):
a stochastic Ensemble Kalman Filter [@Evensen1994; @Burgers1998enkf] that
sequentially assimilates new in-situ observations into the ConvLSTM
forecast without retraining the network.

```{r load, message = FALSE}
library(edaphos)
library(ggplot2)

res_path <- system.file("extdata", "temporal_cerrado_results.rds",
                         package = "edaphos")
if (!nzchar(res_path) || !file.exists(res_path)) {
  knitr::knit_exit(
    "Pre-computed Pillar 3 real-data bundle not found — skipping."
  )
}
R <- readRDS(res_path)
str(R$meta, max.level = 1L)
```

## 2. The data cube

The cube has shape `(H = 10, W = 10, T = 168, C = 3)`. The figure
below shows the three channels for three representative months:
February 2010 (austral mid-wet-season), August 2015 (mid-dry-season),
December 2023 (final forecast target).

```{r show-channels, eval=requireNamespace("ggplot2", quietly = TRUE)}
if (!is.null(R$truth_future)) {
  # The saved cube has past + future separated. Reconstruct a thin
  # sample for plotting: last month of past (Dec 2020, t=132) from the
  # truth_past (not saved slim), fall back to a future slice.
  df_truth <- data.frame(
    row = rep(seq_len(R$meta$H), times = R$meta$W),
    col = rep(seq_len(R$meta$W), each  = R$meta$H),
    ndvi_norm = as.vector(R$truth_future[R$meta$T_future, , ])
  )
  df_truth$ndvi <- R$scaling$ndvi$mu + R$scaling$ndvi$sd * df_truth$ndvi_norm
  ggplot(df_truth, aes(col, -row, fill = ndvi)) +
    geom_raster() +
    scale_fill_viridis_c(name = "NDVI") +
    coord_equal() + theme_minimal(base_size = 12) +
    labs(x = NULL, y = NULL,
         title = sprintf("Observed NDVI over the Cerrado AoI, %s",
                          format(R$meta$target_month, "%b %Y")),
         subtitle = "MOD13Q1 250 m monthly-aggregated mean")
}
```

## 3. Train / forecast split

We train on the first **132 months** (Jan 2010 – Dec 2020) and forecast
the last **36 months** (Jan 2021 – Dec 2023). **Ten** ConvLSTM
ensemble members are fit with different random seeds, sharing the
same architecture (two stacked layers with hidden widths 8 and 4,
3×3 kernels, 120 Adam epochs at learning rate 0.02). A pilot run
with `K_ens = 5` and tight observation noise showed clear ensemble
collapse — the posterior RMSE grew above the prior RMSE because the
undersampled sample covariance over-fitted 20 observations; K = 10
with 8 observations at `obs_sd = 0.15` (≈ 0.026 raw NDVI) is a
stable minimum for this problem.

```{r loss-curves, eval=requireNamespace("ggplot2", quietly = TRUE)}
loss_df <- do.call(rbind, lapply(seq_along(R$training$loss_histories),
                                    function(k) {
  data.frame(epoch  = seq_along(R$training$loss_histories[[k]]),
             loss   = R$training$loss_histories[[k]],
             member = factor(k))
}))
ggplot(loss_df, aes(epoch, loss, colour = member)) +
  geom_line(linewidth = 0.6) + scale_y_log10() +
  theme_minimal(base_size = 12) +
  labs(x = "Epoch", y = "MSE on normalised NDVI (log)",
       title = sprintf("ConvLSTM training loss — %d-member ensemble",
                        R$training$K_ens))
```

The final training losses span
`r sprintf("[%.4f, %.4f]", min(R$training$final_losses), max(R$training$final_losses))`
on normalised NDVI — a narrow spread that confirms the ensemble
sampling is dominated by initialisation noise rather than by genuinely
different attractors.

## 4. The rollout forecast ensemble at the target month

For every ensemble member we re-apply the ConvLSTM to the
concatenated past + future sequence (rollout, see
`?temporal_convlstm_rollout`) and keep the prediction for the last
future month (Dec 2023). The result is a forecast ensemble of shape
`(K = 10, H = 10, W = 10)` — the input to the Kalman update.

```{r prior-map, eval=requireNamespace("ggplot2", quietly = TRUE)}
prior_df <- data.frame(
  row = rep(seq_len(R$meta$H), times = R$meta$W),
  col = rep(seq_len(R$meta$W), each  = R$meta$H),
  ndvi_norm = as.vector(R$prior$mean)
)
prior_df$ndvi <- R$scaling$ndvi$mu + R$scaling$ndvi$sd * prior_df$ndvi_norm
ggplot(prior_df, aes(col, -row, fill = ndvi)) +
  geom_raster() +
  scale_fill_viridis_c(name = "prior NDVI") +
  coord_equal() + theme_minimal(base_size = 12) +
  labs(x = NULL, y = NULL,
       title = sprintf("Prior ensemble-mean NDVI forecast, %s",
                        format(R$meta$target_month, "%b %Y")))
```

Prior RMSE against MOD13Q1 truth:
**`r sprintf("%.4f", R$prior$rmse)`** NDVI z-units
(≈ `r sprintf("%.4f", R$prior$rmse * R$scaling$ndvi$sd)` raw NDVI).

## 5. Sequential Bayesian update with `temporal_kalman_update()`

We simulate the arrival of `r R$analysis$n_obs` fresh in-situ NDVI
observations: the "new campaign" returns truth + Gaussian noise of
standard deviation `r sprintf("%.3f", R$obs$sd)` (normalised units,
i.e. ≈ `r sprintf("%.3f", R$obs$sd * R$scaling$ndvi$sd)` NDVI). The
**stochastic Ensemble Kalman Filter** nudges every member of the
forecast ensemble toward the new evidence through

$$
\mathbf X_a^{(i)} = \mathbf X_f^{(i)} +
                    \bigl(K \odot L\bigr)\,(\mathbf y +
                         \boldsymbol\varepsilon^{(i)} -
                         H\,\mathbf X_f^{(i)}),\qquad
i = 1,\ldots,K_\text{ens},
$$

with the gain
$K = P_f H^{\top}(H P_f H^{\top} + R)^{-1}$ estimated directly from
the ensemble sample covariance [@Evensen1994; @Burgers1998enkf] and
$L$ a Gaspari–Cohn (1999) 5-th-order polynomial taper zeroing the
update beyond `2 * localization_radius` grid cells
[@Houtekamer2001enkf]. The taper is the standard fix for spurious
long-range correlations in small-ensemble EnKFs; with `K = 10` the
un-localized run exhibits ensemble collapse (posterior RMSE above
prior RMSE), while the localized run with `localization_radius = 2`
recovers a net RMSE reduction.

```{r kalman-diag}
cat(sprintf("n_obs    = %d\n", R$analysis$n_obs))
cat(sprintf("n_ens    = %d\n", R$analysis$n_ens))
cat(sprintf("innovation (obs - forecast-at-obs-cell), summary:\n"))
print(summary(R$analysis$innovation))
cat(sprintf("\ngain_row_norm, summary (L2 norm of each obs column of K):\n"))
print(summary(R$analysis$gain_row_norm))
```

### 5.1 Posterior mean and RMSE gain

```{r analysis-map, eval=requireNamespace("ggplot2", quietly = TRUE)}
analysis_df <- data.frame(
  row = rep(seq_len(R$meta$H), times = R$meta$W),
  col = rep(seq_len(R$meta$W), each  = R$meta$H),
  ndvi_norm = as.vector(R$analysis$mean)
)
analysis_df$ndvi <- R$scaling$ndvi$mu + R$scaling$ndvi$sd * analysis_df$ndvi_norm
# Mark the observed cells on the map.
obs_df <- data.frame(
  row  = R$obs$row,
  col  = R$obs$col,
  ndvi = R$scaling$ndvi$mu + R$scaling$ndvi$sd * R$obs$value
)
ggplot(analysis_df, aes(col, -row, fill = ndvi)) +
  geom_raster() +
  geom_point(data = obs_df, aes(col, -row), colour = "white",
             fill = "red", shape = 21, size = 2.6, stroke = 0.5,
             inherit.aes = FALSE) +
  scale_fill_viridis_c(name = "posterior NDVI") +
  coord_equal() + theme_minimal(base_size = 12) +
  labs(x = NULL, y = NULL,
       title = sprintf("Posterior ensemble-mean NDVI, %s",
                        format(R$meta$target_month, "%b %Y")),
       subtitle = sprintf(
         "Red points = %d assimilated in-situ observations",
         R$analysis$n_obs))
```

```{r rmse-gain}
cat(sprintf("Prior     RMSE = %.4f (NDVI z-units)\n", R$prior$rmse))
cat(sprintf("Posterior RMSE = %.4f  (%.1f%% reduction)\n",
            R$analysis$rmse,
            100 * (1 - R$analysis$rmse / R$prior$rmse)))
```

### 5.2 Posterior uncertainty reduction

The Kalman gain shrinks ensemble spread most in the neighbourhood of
each observation. A map of the **normalised uncertainty reduction**
`1 − posterior_sd / prior_sd` shows the spatial footprint of the
ensemble-estimated observation influence function:

```{r sd-reduction-map, eval=requireNamespace("ggplot2", quietly = TRUE)}
sd_df <- data.frame(
  row = rep(seq_len(R$meta$H), times = R$meta$W),
  col = rep(seq_len(R$meta$W), each  = R$meta$H),
  ratio = as.vector(1 - R$analysis$sd / pmax(R$prior$sd, 1e-8))
)
ggplot(sd_df, aes(col, -row, fill = ratio)) +
  geom_raster() +
  geom_point(data = obs_df, aes(col, -row), colour = "black",
             fill = "white", shape = 21, size = 2.6, stroke = 0.5,
             inherit.aes = FALSE) +
  scale_fill_viridis_c(option = "magma", name = "1 − σ_a / σ_f",
                        limits = c(0, 1)) +
  coord_equal() + theme_minimal(base_size = 12) +
  labs(x = NULL, y = NULL,
       title = "Posterior uncertainty reduction",
       subtitle = sprintf(
         "Mean σ_a / σ_f = %.3f  (lower ⇒ more information gained)",
         R$sd_reduction))
```

## 6. What this vignette buys you over the v1.3 Pillar 3

1. **Real remote-sensing / climate data**, not
   `temporal_synth_soc_cube()` synthetic. 168 real monthly time steps
   over a 400-cell Cerrado AoI, three published sources with DOIs.
2. **A multi-step rollout forecast** against the last three years of
   MOD13Q1 NDVI, reported as RMSE in both normalised and raw NDVI
   units.
3. **A Bayesian assimilation step** that folds new ground observations
   into a trained ConvLSTM without retraining — the stochastic EnKF
   of @Evensen1994 adapted to the ensemble produced by K independent
   seeds of `temporal_convlstm_fit()`.
4. **Quantitative diagnostics** of information gain: innovation
   vector, per-observation gain norm, and the map of
   `1 − σ_a / σ_f` that makes the spatial footprint of each
   observation explicit.

### 6.1 Caveats and planned extensions

* The ensemble is sampled *only* by the random seed of the optimiser.
  Dropout ensembles, MC-dropout heads and deep-ensemble hyperparameter
  perturbation [@Lakshminarayanan2017deepens] are alternative
  spread-generating mechanisms, scheduled for the unified-uncertainty
  pillar (v1.6).
* The Gaspari–Cohn [@Gaspari1999] localization radius is a fixed
  hyperparameter; adaptive localization schemes that choose the taper
  bandwidth per observation from the ensemble spread itself are the
  obvious next refinement.
* The cube covers only 10 × 10 cells at 0.2° resolution. Scaling to
  the operational 1 km² resolution over the full Cerrado
  (~2 million cells) is a pure compute question; the algorithmic
  machinery presented here is unchanged.

## References
